```{r}
#| echo: false
#| cache: false
source("_common.R")
```

# Survival analysis {#sec-survival}


::: {.callout-warning}
This chapter is still a draft and minor changes are made without notice.
In particular, a few practical examples will be added in Sections 11.4 and 11.5.
:::

This chapter deals with regression models for time-to-event data with possible right censoring, which 
is common for survival data but also in other applications of statistics[^app] such as reliability 
theory and credit risk. 

[^app]: Time to failure of an electronic component, say, or time default of a debitor.

Some of the notation in this chapter will deviate a bit from 
the notation used for other regression models, which is deliberate and to follow 
some of the standard conventions in the survival literature. 
The response variable will, for example, be denoted $T$ instead of $Y$, which is 
also a natural notation for a time-to-event observation. 
The regression models are, however, still of the form, where a linear 
predictor $\eta = X^T \beta$ is formed from the vector of predictors $X \in \mathbb{R}^p$ 
and a parameter $\beta \in \mathbb{R}^p$, and the effect of the predictors on 
the survival distribution is then only through $\eta$. 



## Outline and prerequisites 

@sec-intro-surv is covers the basic definitions specific to survival analysis 
and motivates why survival (regression) analysis is special. It also covers the
basic nonparametric estimators of survival distributions. @sec-par-surv 
covers parametric survival models and likehood based methods for estimation 
and statistical inference. The accelerated failure time models and the proportional
hazard model are introduced, which show two different ways that the linear 
predictor can affect the survival distribution. @sec-semipar-surv primarily 
treats Cox's proportional hazard model, which is the most widely used 
survival regression model. As the name suggests, it is a proportional hazard model, 
but the methodology is *semiparametric*, meaning in this case that a baseline hazard
function is left unconstrained. @sec-res-surv introduces survival residuals, 
which differ a bit from residuals considered for linear and generalized linear models. 
These differences reflect that survival models don't target conditional expectations, 
but otherwise the purposes of the residuals for model diagnostics are the same. 

The chapter primarily relies on the following R packages. 

```{r}
#| label: setup
#| cache: false
#| code-fold: false
#| message: false
library(RwR)
library(ggplot2)
library(tibble)
library(dplyr)
library(tidyr)
library(skimr)
library(broom)
library(lubridate)
library(survival)
library(splines)
```


## Introduction to survival analysis {#sec-intro-surv}

A typical application of survival analysis is to the study of survival of patients 
after an initial diagnosis. In such a study, patients are enrolled whenever
they are diagnosed with a given (serious, life threatening) disease. 
We refer to the time of enrollment[^baseline] as the *baseline*, and we note that 
the baseline for different patients can be different 
calendar times, and that the survival time is the time from baseline to death.
The baseline is thus time zero on the time axis of the survival time.

[^baseline]: For some applications, all individuals or units can have the same baseline, e.g., in an experiment on the longevity of electronic components, but in many applications the baseline calendar times differ among units. In credit risk models of time to default, say, the baseline could be the date of the loan application. 

The patients are followed for some period of time, and if they 
die during this followup period, their survival time is registered. 
At a planned calendar time the study ends and the status of the patients are registered, that is, 
we register if they died, in which case we know their survival time, or if they 
are still alive. 
All patients alive at the end of the study have *right censored* survival 
times. This form of censoring is sometimes called *administrative censoring*. 
Some patients might also have dropped out of the study earlier, because they moved or didn't want to 
participate anymore, say. At the time they drop out, we say that they are *lost-to-followup*, 
and this also results in right censoring of their survival time. 

In addition to the, possibly right censored, survival times of the patients, 
we may collect additional data on the patients. We can then use some or all 
of such additional variables to construct a regression model of the 
survival distributions. But it is not clear that 
we want to model *expected* survival times, and the conditional expectation 
is often not[^reasons] the correct target of a survival regression model. 

[^reasons]: One reason is technical. Expectations rely on the tails of distributions, and with right censoring we don't get that information. 

Regression models for survival data 
thus differ from the linear and generalized linear regression 
models in at least the following two ways:

* There is almost always a *censoring mechanism*, and certain aspects of the
    data are consequently missing. We need to deal with this in the modeling.
* Survival distributions are skewed distributions on the positive half line. 
    It is often the *shape* of the distribution rather than the location of the distribution
    that is of interest.

The overall objective of a survival regression model is the same as for any regression model: 
how are the different predictors associated with the survival time. We might then have 
different more specific questions we want to answer. For example, to determine if the survival distributions 
depends on a choice of treatment (testing). Or we might want to quantify how strongly 
a predictor is associated with survival response.

We might, of course, also be interested in prediction, which we often call *survival prognosis*. 
It is rarely of interest to make a point prediction of a survival time for an individual, 
but we might report a prediction interval (the interquartile range, say), or we might report 
probabilistic predictions in terms of an entire (conditional) survival distribution.

::: {#exm-prostate}

We will throughout this chapter use a dataset on prostate cancer survival, 
which is included in the RwR package as the dataset `prostate`.
The data is from a randomized clinical trial on prostate
cancer patients with four different treatments (one placebo and three
different dosages of estrogen). There are 502 patients in the trial,
and we will in this chapter focus on four of the variables.

* `rx`: Treatment encoded as a factor.
* `status`: Either the cause of death or indication that the patient was alive at the end of study.
* `dtime`: The censored survival time since baseline in months.
* `age`: The baseline age of the patient in years. 

There are several other variables included in the dataset, and among 
these are `sdate`: the date that the patient entered the study. 
We will not use `sdate` for modeling, but it is used in @fig-prostate-entry 
to visualize the data. For later purposes, the data is converted to a tibble.

```{r}
#| label: prostate-data
data(prostate, package = "RwR") # version >= 0.3.5
prostate <- prostate |> as_tibble()
```

```{r}
#| label: fig-prostate-entry
#| fig-cap: "The entry date of the 100 first prostate cancer patients into the study, their follow-up time and their status at end of follow-up."
#| column: margin
#| fig-height: 7
#| fig-width: 3
#| echo: false
head(prostate, 100) |> 
    arrange(sdate) |>
    mutate(index = seq_along(sdate)) |> 
    ggplot(aes(x = sdate, xend = sdate + months(dtime), y = index)) +
    geom_segment() + 
    geom_point(aes(x = sdate + months(dtime), y = index, color = (status != "alive"))) +
    scale_color_brewer("status:", labels = c("censored", "dead"), palette = "Set1") +
    ylab("Patient") +
    xlab("Date") +
    theme(legend.position = "top")
select(prostate, c(rx, status, dtime, age, sdate)) 
```

@fig-prostate-entry illustrates the data for the first 100 patients 
that enter the study. The figure shows their entry date (baseline), 
their survival time from that date and the status indicator when follow-up ends, 
indicating if they are dead or if the survival time is censored.   

Since the treatments were randomly assigned, differences 
in the survival distributions between the four treatment groups 
can be ascribed to the treatment alone. That is, the treatments 
can be interpreted as the *causes* of any such differences. We
will first of all be interested in testing if there are effects 
of treatment as well as estimating the magnitudes of such effects. 

There are other variables included in the dataset, such as 
age, and we could use these variables for two different purposes. 
We could use the variables for prognostic purposes, that is, for 
fitting regression models that give more accurate survival prognoses 
for individual patients. Another purpose could be to investigate 
*heterogeneity* of treatment effects, that is, to investigate 
if the treatment works differently on different subgroups of the 
patients. 
:::

### Survival distributions and censoring

A survival time $T^*$ is simply a non-negative random variable. We will denote 
its distribution function by $F$, and if the distribution has density $f$ w.r.t. Lebesgue measure, 
we have 
$$
    F(t) = \mathbf{P}(T^* \leq t) = \int_{0}^{t} f(s) \mathrm{d}s.
$$ 
The corresponding *survival function* of $T^*$ is 
$$
    S(t) \coloneqq 1 - F(t) = \mathbf{P}(T^* > t) = \int_t^{\infty} f(s) \mathrm{d} s.
$$ 

In terms of $S$ we define the *cumulative hazard function* as 
$$
    \Lambda(t) \coloneqq - \log(S(t)),
$$ {#eq-cum-def}
so that 
$$
    S(t) = \exp\left(-\Lambda(t)\right).
$$ {#eq-surv-from-cum} 
If $S$ is continuously differentiable, $-S'(t) = f(t)$ is the density of the survival distribution, 
and the *hazard function* is defined as 
$$
    \lambda(t) \coloneqq - (\log S(t))' = \frac{f(t)}{S(t)}.
$$ {#eq-hazard}
[There are one-to-one relations between the distribution function, the survival function and the cumulative hazard function.]{.aside}

We can observe that 
$$ 
    \begin{align*}
        \lambda(t) & = \lim_{\varepsilon \to 0+} \frac{1}{\varepsilon} \frac{F(t+\varepsilon) - F(t)}{S(t)} \\
        & =  \lim_{\varepsilon \to 0+} \frac{1}{\varepsilon} \mathbf{P}(T^* \in (t,t+\varepsilon] \, | \, T^* > t),
    \end{align*}
$$
and $\lambda(t)$ thus has the interpretation as the instantaneous rate of death at time $t$ given survival beyond time $t$. 
We also note that
$$
    \Lambda(t) = \int_0^t \lambda(s) \mathrm{d} s = - \log S(t)
$$
which explains why $\Lambda$ is called the *cumulative* hazard function.

We will not observe $T^*$ but only 
$$
    T = \min\{T^*, C\} \quad \text{ and } \quad \delta = 1(T^* \leq C)
$$
for a *censoring variable* $C$. We call $T$ the *right censored* survival time 
and $\delta$ the *status indicator*. The status indicator is $1$ if the $T$ is uncensored and 
$0$ otherwise. To distinguish the distribution of $C$ clearly from the distribution of the survival 
time, we will generally denote its survival function by $H$ and its density by $h$ (if
it exists). That is, 
$$
    H(t) = \mathbf{P}(C > t) = \int_t^\infty h(s) \mathrm{d} s.
$$

### Model assumptions

We will generally consider $n$ observations and let $T^*_1, \ldots, T^*_n$ denote
the survival times and $C_1,\ldots,C_n$ the censoring times. We then observe
the right censored survival times 
$$
    T_i = \min\{T_i^*, C_i\}
$$
and the status indicators
$$
    \delta_i = 1(T_i^* \leq C_i)
$$
for $i = 1, \ldots, n$. 

The vector of predictor variables is as elsewhere denoted $X_i \in \mathbb{R}^p$, and we are generally 
interested in models of the conditional distributions[^cond-dist] 
$$
    T_i^* \mid X_i = x.
$$ 

An issue of particular concern for survival analysis is *when* a predictor variable is observed. In some studies, 
some potential predictor variables are monitored throughout the study. Such variables as BMI and 
blood pressure might be recorded on a monthly basis, say. It is possible to develop a 
general time-to-event theory where predictors also change with time, but it is both technically 
beyond this chapter, and it is also conceptually challenging to understand and interpret 
such models correctly.

[^cond-dist]: Note that the target is the conditional distribution of the unobserved and uncensored $T^*_i$, not the oberved and censored $T_i$. 

We will throughout assume that all predictors are observed[^observable] at or 
before baseline, and we refer to such predictors as *baseline predictors*.

The modeling assumptions across the different individuals are now straightforward 
modifications of A5 or A6 to the survival setup.

[^observable]: Or, as a minimum, observable at time zero.

**SA5:** The conditional distribution of $(T^*_i, C_i)$ given $\mathbf{X}$ depends only 
on $X_i$ and $(T^*_1, C_1), \ldots, (T^*_n, C_n)$ are conditionally independent given $\mathbf{X}$.

**SA6:** The triples $(T^*_1, C_1, X_1), \ldots, (T^*_n, C_n, X_n)$ are i.i.d. 

Since $(T_i, \delta_i)$ is a function of $(T^*_i, C_i)$, then either assumption 
implies the same distributional statement about the $(T_i, \delta_i)$-s as about 
the $(T^*_i, C_i)$-s.

An additional delicate survival modeling question is what to assume about the 
joint distribution of $T^*_i$ and $C_i$ (possibly, conditionally on $X_i$). 
Throughout, we will make the following assumption.

**Independent censoring:** $T^*_i$ and $C_i$ are conditionally independent given $X_i$. 

Even though the assumption above is a *conditional* independence statement, 
it is known simply as "independent censoring". In the literature, 
"independent censoring" may refer to a slightly more permissive 
assumption (Definition III.2.1 in @AndersenBorganGillKeiding:1993), 
stating that information about censoring status at any given timepoint 
is not allowed to affect the residual survival distribution beyond that 
timepoint. For this chapter we will make the stronger, but simpler, 
independent censoring assumption as stated above. 

### The Kaplan-Meier and Nelson-Aalen estimators

If $T_1^*, \ldots, T_n^*$ are i.i.d. and if they were observed, we could define the empirical 
survival function 
$$
    \hat{S}^*_n(t) = \frac{1}{n} \sum_{i=1}^n 1(T_i^* > t)
$$
as a nonparametric estimator of $S$. Clearly, we cannot compute 
$\hat{S}^*_n$ if we only have censored observations. 
The purpose of this section is to 
introduce two nonparametric estimators of a survival distribution 
when the survival times are i.i.d. and independently censored[^ind-cens]. 

[^ind-cens]: In this section there are no predictors, and we effectively assume that $(T_1^*, C_1), \ldots, (T_n^*, C_n)$
are i.i.d. and that $T_i^*$ and $C_i$ are independent. 

To handle the censoring we introduce the *individuals at risk* at time $t$:
$$
    Y(t) = \sum_{i=1}^n 1(t  \leq T_i).
$$
It can be computed based on the censored survival times, and it counts 
how many individuals have survived to time $t$ *and are uncensored* at 
that time.  

@def-KM below gives the general definition of the *Kaplan-Meier* estimator 
of the survival function, but to motivate this definition, we
first suppose that there are no ties among the observed survival times, 
that is, where we have strict inequalities among the ordered survival times
$$
    T_{(1)} < T_{(2)} < \ldots < T_{(n-1)}  < T_{(n)}.
$$
Here $T_{(k)}$ denotes the survival time with rank $k$, and we will 
denote the corresponding status indicator by $\delta_{(i)}$. 
[Without censoring, the Kaplan-Meier estimator equals $\hat{S}^*$, cf. @exr-empsurv]{.aside}

When there are no ties, and $T_{(k)} \leq t < T_{(k+1)}$, we can factorize the survival function as 
$$
    \begin{align*}
        S(t) & =  \mathbf{P}(T^* > t) \\
        & =  \mathbf{P}(T^* > t \mid T^* \geq T_{(k)}) \times  \mathbf{P}(T^* \geq T_{(k)} \mid T^* \geq T_{(k-1)}) \\
        & \hspace{5mm} \ldots \times    \mathbf{P}(T^* \geq T_{(2)} \mid T^* \geq T_{(1)}) \times \mathbf{P}(T^* \geq T_{(1)}) \\
    \end{align*}
$$ 
where $\mathbf{P}(T^* \geq T_{(i+1)} \mid T^* \geq T_{(i)})$ is the conditional probability of surviving 
from $T_{(i)}$ to $T_{(i+1)}$ given survival beyond time $T_{(i)}$. 

Since $Y(T_{(i)})$ are the number of individuals at risk[^at-risk] at time $T_{(i)}$ 
and $Y(T_{(i)}) - \delta_{(i)}$ are the number of individuals that die in the interval 
$[T_{(i)}, T_{(i+1)})$, we can regard 
$$
    \frac{Y(T_{(i)}) - \delta_{(i)}}{Y(T_{(i)})} = 1-\frac{\delta_{(i)}}{Y(T_{(i)})}
$$
as an estimator of the conditional probability $\mathbf{P}(T^* \geq T_{(i+1)} \mid T^* \geq T_{(i)})$.
This gives the *Kaplan-Meier* estimator for $t \in [T_{(k)}, T_{(k+1)})$
$$
    \hat{S}(t) = \left(1-\frac{\delta_{(k)}}{Y(T_{(k)})}\right) 
    \cdots 
    \left(1-\frac{\delta_{(2)}}{Y(T_{(2)})}\right) 
   \left(1-\frac{\delta_{(1)}}{Y(T_{(1)})}\right).
$$ 

[^at-risk]: Alive and *uncensored*.

We can write the Kaplan-Meier estimator in a slightly more abstract notation as 
$$
    \hat{S}(t) = \prod_{i: T_i \leq  t} \left(1-\frac{\delta_i}{Y(T_i)}\right).
$$ {#eq-KM-no-ties}


To define the Kaplan-Meier estimator that allows for multiple deaths 
at the same time (ties), we introduce  the *counting process* of deaths (non-censored events)
$$N(t) = \sum_{i=1}^n 1(T_i \leq t, \delta_i=1),$$
and the jumps for the counting process given as 
$$\Delta N(t) = N(t) - N(t-) = N(t) - \lim_{\varepsilon \to 0+} N(t-\varepsilon),$$

\index{Kaplan-Meier estimator} 

::: {#def-KM}
With the above notation, the Kaplan-Meier estimator is defined as
$$
    \hat{S}(t) = \prod_{s \leq t} \left(1-\frac{\Delta N(s)}{Y(s)}\right).
$$ {#eq-KM-general}
:::

In the infinite product above, all factors where $\Delta N(s) = 0$ 
are equal to $1$. There is thus only a finite number of factors different from $1$. 
When $\Delta N(s) \in \{0, 1\}$ we see that @eq-KM-general coincides with 
@eq-KM-no-ties. The definition allows for $\Delta N(s) > 1$
to correctly account for multiple deaths at the same time and thus ties among 
the observed survival times. 
[The function `survfit()` from the R package survival can be used to compute the Kaplan-Meier estimator.]{.aside}


Since the cumulative hazard function is $\Lambda = - \log(S)$, we can estimate it 
nonparametrically as 
$$
    - \log(\hat{S}(t)) = - \sum_{s \leq t} \log  \left(1-\frac{\Delta N(s)}{Y(s)}\right).
$$ {#eq-cum-by-KM}
But there is also a different, and direct, nonparametric estimator of the cumulative hazard function.
The logic behind this estimator is that for a *differentiable*
survival function $S$ and $t > s$ (but $t - s$ small),

$$
    \begin{align*}
        \mathbf{P}(T^* \in [s, t) \mid T^* \geq s) & = \frac{S(s) - S(t)}{S(s)} \\
        & \approx - \frac{S'(s)}{S(s)}(t - s) = \lambda(s)(t - s).
    \end{align*}
$$

By a Riemann sum approximation we see that 
$$
    \Lambda(t) \approx \sum_{i: T_{(i)} \leq t}  \mathbf{P}(T^* \in [T_{(i)}, T_{(i + 1)}) \mid T^* \geq T_{(i)}), 
$$
and if there are no ties, the conditional probabilities in the sum above can be estimated[^est-prob] as 
$\delta_{(i)} / Y(T_{(i)})$. This leads to the *Nelson-Aalen* estimator 
$$
    \hat{\Lambda}(t) = \sum_{i: T_{(i)} \leq t} \frac{\delta_{(i)}}{Y(T_{(i)})}.
$$ {#eq-NA-no-ties}

[^est-prob]: Either $\delta_{(i)} = 1$ and one individual out of $Y(T_{(i)})$ dies in the interval (at time $T_{(i)}$), or $\delta_{(i)} = 0$ and no individuals die in the interval.

The general definition below allows for multiple deaths at the same time.

::: {#def-NA}
The Nelson-Aalen \index{Nelson-Aalen estimator} estimator of the cumulative hazard function is  
$$
    \hat{\Lambda}(t) = \sum_{s \leq t}  \frac{\Delta N(s)}{Y(s)}.
$$ {#eq-NA-general}
:::

Since the survival function is given by $S(t) = \exp(-\Lambda(t))$, we can
estimate the survival function by plugging the 
Nelson-Aalen estimator of the cumulative hazard, which gives
$$
    \exp\left(-\hat{\Lambda}(t)\right) = \prod_{s \leq t} \exp\left(-
    \frac{\Delta N(s)}{Y(s)}\right).
$$ {#eq-surv-by-NA}

We note that $\exp(-\hat{\Lambda})$ is not the Kaplan-Meier estimator and nor is 
$- \log(\hat{S})$ the Nelson-Aalen estimator. 
[The function `survfit()` with argument `stype = 2` estimates the survival function in terms of the Nelson-Aalen estimator via @eq-surv-by-NA.]{.aside}

The mismatch between the Kaplan-Meier and the Nelson-Aalen estimators is a small nuisance, 
but we may note that 
$$ 
    \exp\left(- \frac{\Delta N(s)}{Y(s)}\right) \approx  \left(1 -
    \frac{\Delta N(s)}{Y(s)}\right)
$$
for large $Y(s)$ by using the Taylor expansion $\exp(x) = 1 + x + o(x^2)$. 
Thus differences between the factors are small unless $s$ is so large that 
the number of individuals at risk, $Y(s)$, is small. 

One can argue that the way ties are handled above presumes that survival times 
can be exactly equal, which is at odds with any continuous survival distribution. 
If we regard ties as a consequence of lack of precision in our registration or due to 
rounding of the survival times, it makes sense to replace $\frac{\Delta N(s)}{Y(s)}$
by 
$$
    \frac{1}{Y(s)} + \frac{1}{Y(s) - 1} + \ldots + \frac{1}{Y(s) - \Delta N(s) + 1}.
$$ 
[The function `survfit()` with argument `ctype = 2` will use the Fleming-Harrington correction for ties.]{.aside}
This is known as the *Fleming-Harrington* correction for ties. It often doesn't
make much of a practical difference though. 

::: {#exm-prostate-II}

We continue @exm-prostate by first considering a summary of the marginal distributions 
of the four variables. 

```{r}
#| label: tbl-skim-prostate
#| skimr_include_summary: FALSE
#| tbl-cap: "Data summary of the four variables in the `prostate` dataset."
#| tbl-cap-location: bottom
#| column: page-right
select(prostate, c(rx, status, dtime, age)) |> skim()
```

@tbl-skim-prostate shows 148 patients (out of 502) were alive at the end of the 
study. Thus 148 of the survival times are right censored. @tbl-prostate-death 
shows a more detailed tabulation of the values of the `status` variable, which 
specifies one out of nine different causes of death -- in addition to specifying 
if the patient is still alive. We may also note that the survival time (`dtime`)
is registered in months since baseline, which is a very coarse scale. Consequently 
there are many ties in the data. 

```{r}
#| label: tbl-prostate-death
#| column: margin
#| tbl-cap: "Tabulation of different causes of death."
#| tbl-cap-location: bottom
#| echo: false
prostate[, "status", drop = FALSE] |> 
    table() |>
    knitr::kable()
```

We will fit a survival distribution, using either of the nonparametric estimators 
above, where we ignore the information about different causes of death. Thus 
the `status` variable will only be used to determine the censoring status.

We fit survival functions for all patients using the `survfit()` function 
from the survival package. See `?survfit.formula` for details about how to 
specify a model using the formula interface. Note that the left hand side of the 
formula, that specifies the response, uses the special `Surv()` function syntax 
to include information about the censoring status. We fit the survival functions 
below using both the Kaplan-Meier and the Nelson-Aalen, and we also fit the 
survival function using the Nelson-Aalen estimator with the Fleming-Harrington 
correction for ties. 

```{r}
#| label: prostate-survfit
#| code-fold: show
prostate_KM <- survfit(
    Surv(dtime, status != "alive") ~ 1, 
    data = prostate,
)

prostate_NA <- survfit(
    Surv(dtime, status != "alive") ~ 1, 
    data = prostate,
    stype = 2
)

prostate_NA_FH <- survfit(
    Surv(dtime, status != "alive") ~ 1, 
    data = prostate,
    stype = 2,
    ctype = 2
)
```

@fig-survival-est shows the fitted survival curves -- including a plot that zooms in on 
a part of the curves. The differences are small and barely visible, and they are most pronounced 
in the tail of the distribution. 

```{r}
#| label: fig-survival-est
#| fig-cap: "Left: estimated survival function for patients with prostate cancer using the Kaplan-Meier estimator (black), the Nelson-Aalen estimator (blue) and the Nelson-Aalen estimator with the Fleming-Harrington correction for ties (red). Right: the same as the left figure but zoomed in."
#| fig-cap-location: margin
#| opts.label: widefigure2
#| code-fold: true
p1 <- summary(prostate_KM, data.frame = TRUE) |> 
        ggplot(aes(time, surv)) + 
        geom_step(data = summary(prostate_NA, data.frame = TRUE), color = "blue") +
        geom_step(data = summary(prostate_NA_FH, data.frame = TRUE), color = "red") +
        geom_step() +
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability")

p2 <- p1 + coord_cartesian(xlim = c(50, 75), ylim = c(0.20, 0.40))

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

We then fit a separate survival function for each treatment group. We can think about this as 
a simple regression model with a single categorical predictor variable (the treatment variable `rx`), 
and the formula specification below is similar to other regression model specifications. The 
main difference is that using `survfit()` we fit a survival function for each treatment group instead of a 
mean value, say. 

```{r}
#| label: prostate-surv-rx
#| code-fold: show
prostate_KM_rx <- survfit(
    Surv(dtime, status != "alive") ~ rx, 
    data = prostate
)
```

```{r}
#| label: fig-prostate-surv-rx
#| fig-cap: "Kaplan-Meier estimators of the survival function for each of the four treatment groups."
#| fig-cap-location: margin
summary(prostate_KM_rx, data.frame = TRUE) |> 
        ggplot(aes(time, surv, color = strata)) + 
        geom_step() +
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top")
```
::: 

## Parametric survival models {#sec-par-surv}

In this section we focus on parametric models of the survival distribution, 
and we will first derive a generic expression for the likelihood function. 
We then turn to regression models and introduce the accelerated failure time 
models and the proportional hazards models as two different was for the 
predictors to affect the response distribution.

### The likelihood

Recall that in terms of $T^*$ and $C$ we have $T = \min\{T^*, C\}$ and $\delta = 1(T^* \leq C)$. 
Our first result gives the density of the joint distribution of $(T, \delta)$ in terms of 
the marginal distributions of $T^*$ and $C$ when they are independent.

::: {#thm-density-surv}
If $T^*$ and $C$ are independent the joint distribution of $(T, \delta)$ has density 
$$
    g(t, \delta)  = f(t)^\delta S(t)^{1-\delta} h(t)^{1-\delta} H(t)^\delta
$$
w.r.t. the product measure $m \otimes \tau$ (Lebesgue measure times the counting measure). 
:::

::: {.proof}
First note that
$$
    \begin{align*}
    \mathbf{P}(T \leq t, \delta = 1) & = \mathbf{P}(T^* \leq t, \delta = 1) \\
    & = \mathbf{P}(T^* \leq t,C \geq T^*) \\
    & = \int_0^t f(s) \int_s^{\infty} h(u) \mathrm{d} u \mathrm{d} s \\
    & = \int_0^t f(s) H(s) \mathrm{d} s.
    \end{align*}
$$
Likewise, 
$$
    \mathbf{P}(T \leq t, \delta = 0) = \int_0^t h(s) S(s) \mathrm{d} s,
$$
and we conclude that the density is
$$
    g(t, \delta) = \left\{ \begin{array}{ll}
    f(t) H(t) & \text{if } \delta = 1 \\
    h(t) S(t) & \text{if } \delta = 0.
    \end{array} \right.
$$
:::

Based on the density result above we can introduce the full likelihood for right censored 
observations. With $(T_1, \delta_1), \ldots, (T_n, \delta_n)$ independent the full likelihood is 
\index{likelihood!survival model}
$$
    L = \prod_{i=1}^n  f_i(T_i)^{\delta_i} S_i(T_i)^{1-{\delta_i}} h_i(T_i)^{1-{\delta_i}} H_i(T_i)^{\delta_i}.
$$

For a parametrized model of the distributions of $T^*_1, \ldots, T^*_n$ we have that 
$f_i = f_{i, \beta}$ for some parameter $\beta$. The censoring distributions, given 
by the densities $g_i$, usually hold no information about the parameter $\beta$. 
This implies that 
$$
    L(\beta) = \prod_{i=1}^n f_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i)^{1-{\delta_i}} K_i
$$
with $K_i$ depending on the observations but not the parameter $\beta$.

Note how the derivation of the likelihood makes it explicit what the dominating measure is,
and thus that the joint distribution in fact has a density. The derivation also makes
it clear how the distribution of the censoring mechanism enters, and 
why it can be ignored when it does not depend on the unknown parameter
$\beta$. The argument for ignorance is not independent censoring[^ind-cen]
but that the $g_i$-s do not depend upon $\beta$. Technically, the censoring 
distribution and the parameter $\beta$ are *variation independent*.

[^ind-cen]: Though independent censoring does play a role here for our particular derivation and form of the full likelihood.

By dropping the factors $K_i$, that do not depend upon the parameter $\beta$, 
in the expression above for the likelihood, we arrive at the following 
definition of the survival likelihood. 

::: {#def-likelihood}
With the above setup, the survival likelihood is
$$ 
L(\beta)  =  \prod_{i=1}^n f_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i)^{1-{\delta_i}}.
$$ {#eq-survlike}
:::

It is possible to take a slightly different point of view and
condition on the censoring variables instead. Then the dominating measure
for the $i$-th censored observation, $T_i$, becomes $m + \delta_{C_i}$ 
($\delta_{C_i}$ is the Dirac measure in $C_i$). One arrives at
the same likelihood as above though. 

We can also express the likelihood in terms of the hazard and cumulative 
hazard functions.

::: {#prp-likelihood}
If $f_{i, \beta}$ has hazard function $\lambda_{i,\beta}$ and cumulative hazard function 
$\Lambda_{i, \beta}$, the survival likelihood equals 
$$ 
    L(\beta) = \prod_{i=1}^n \lambda_{i, \beta}(T_i)^{\delta_i} e^{- \Lambda_{i, \beta}(T_i)} 
$$ {#eq-survlike2}
::: 

::: {.proof}
Recall that 
$$
    \lambda_{i, \beta}(t) = \frac{f_{i,\beta}(t)}{S_{i,\beta}(t)}  
$$
and that 
$$
    S_{i,\beta}(t) = e^{- \Lambda_{i, \beta}(t)}.
$$
Hence 
$$
    \begin{align*}
    L(\beta) & = \prod_{i=1}^n f_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i)^{1-{\delta_i}} \\
    & =  \prod_{i=1}^n \lambda_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i)^{1-{\delta_i}} \\
    & = \prod_{i=1}^n \lambda_{i, \beta}(T_i)^{\delta_i} S_{i, \beta}(T_i) \\
    & =  \prod_{i=1}^n \lambda_{i, \beta}(T_i)^{\delta_i} e^{- \Lambda_{i, \beta}(T_i)}.
    \end{align*}
$$
:::

:::{#exm-weibull-hazard}
The *Weibull distribution* has hazard function 
$$
    \lambda(t) = \alpha \gamma t^{\gamma - 1}
$$
for $\alpha, \gamma > 0$ two parameters. The special 
case $\gamma = 1$ gives us the exponential distribution 
with constant hazard function. The Weibull distribution 
has increasing hazard functions over time when $\gamma > 1$
and decreasing hazard function when $\gamma < 1$.

The cumulative hazard function is 
$$
    \Lambda(t) = \alpha  \int_0^t \gamma s^{\gamma - 1} \mathrm{d} s = \alpha t^\gamma.
$$

For right censored i.i.d. observations from the Weibull model, 
the likelihood becomes 
$$
    L(\alpha, \gamma) = \prod_{i=1}^n (\alpha \gamma)^{\delta_i} T_i^{\delta_i(\gamma - 1)} e^{- \alpha T_i^\gamma},
$$
and the log-likelihood is 
$$
    \begin{align*}
    \ell(\alpha, \gamma) & = \log L(\alpha, \gamma) \\
    & = \sum_{i=1}^n \delta_i (\log(\alpha) + \log(\gamma) + (\gamma - 1)\log(T_i)) 
    - \alpha T_i^\gamma.
    \end{align*}
$$
:::

:::{#exm-exponential-MLE}

## MLE for the exponential distribution

Continuing @exm-weibull-hazard and fixing $\gamma = 1$, corresponding to an
exponential distribution with hazard rate $\alpha$, we find that the 
log-likelihood equals 
$$
    \ell(\alpha) = \sum_{i=1}^n \delta_i \log(\alpha) - \alpha T_i
$$
up to additive terms that do not depend on $\alpha$. This log-likelihood has 
a unique stationary point 
$$
    \hat{\alpha} = \frac{\sum_{i=1}^n \delta_i}{\sum_{i=1}^n T_i} = \frac{n_u}{\sum_{i=1}^n T_i}
$$
where $n_u = \sum_{i=1}^n \delta_i$ are the number of deaths (the number of uncensored events). 
The second derivative of of $\ell$ is negative, so $\ell$ is concave and $\hat{\alpha}$ is 
the MLE. 

If we had treated the observations as if there were no censoring, we would get the MLE 
$$
    \frac{n}{\sum_{i=1}^n T_i} > \hat{\alpha},
$$
which overestimates the rate. 

If we, on the other hand, discard the censored observations, we would get the MLE 
$$
    \frac{n_u}{\sum_{i=1}^n \delta_i T_i} > \hat{\alpha},
$$
which likewise overestimates the rate.
:::

### Regression models

We introduce two ways that predictor variables 
can affect the survival response distribution. The accelerated failure time 
models exemplified by the log-logistic model, and the proportional 
hazards models exemplified by the Weibull model. In the latter case, 
it turns out that the likelihood can be computed and optimized 
using methods for Poisson generalized linear models. 

#### Accelerated failure time models {.unnumbered}

::: {#exm-log-logistic}

## Log-logistic distribution

Suppose $Z$ has a logistic distribution with distribution function \index{log-logistic distribution}
$$
    G(z) = \frac{e^{\lambda z}}{1 + e^{\lambda z}}.
$$
The distribution of $T^* = e^Z$ is then the *log-logistic* distribution, whose 
distribution function is 
$$
    F(t) = G(\log(t)) = \frac{t^{\lambda}}{1 + t^{\lambda}},
$$
with corresponding density
$$
    f(t) = F'(t) = \frac{\lambda t^{\lambda-1}}{(1+t^{\lambda})^2}
$$
for $t > 0$.  
:::


::: {#def-AFT} 
Let $G$ be a distribution function (on $\mathbb{R}$). The corresponding accelerated failure time (AFT) model 
has survival function given as 
$$
    S_{\eta} (t) = 1 - G( (\log(t) - \eta)/\sigma)
$$
with $\eta = X^T \beta$ the linear predictor and $\sigma > 0$ a scale parameter. 
[The AFT model is a scale-location model of the log-survival time.]{.aside}
:::

The interpretation of the AFT model is that a 
unit change of the $j$-th predictor increases -- or *accelerates* -- the
failure time by a factor $e^{\beta_j}$. 


::: {#exm-log-logistic-II}

## Log-logistic AFT model

If $G$ is the logistic distribution with scale parameter is $\sigma = \lambda^{-1}$,
the survival function for the corresponding AFT model is 
$$
        \begin{aligned}
            S_{\eta}(t)  & = \frac{1}{1 + e^{\lambda(\log(t) - \eta)}} \\
            & = \frac{1}{1 + e^{-\lambda \eta} t^{\lambda}}.
        \end{aligned}
$$ {#eq-log-logistic-surv}
where $\eta = X^T \beta$ is the linear predictor. This is 
a log-logistic distribution as introduced in @exm-log-logistic. The 
effect of $\eta$ is[^scaling] a *scale transformation* of the log-logistic 
distribution with scale parameter $e^{\eta}$. The density is 
$$
    \begin{aligned}
        f_{\eta}(t) & = \frac{\lambda e^{-\lambda \eta} t^{\lambda-1}}{(1+e^{-\lambda \eta} t^{\lambda})^2} \\
        & = \frac{\lambda e^{\lambda (\log(t) - \eta)} t^{-1}}{(1+e^{\lambda(\log(t)  - \eta)})^2}.
    \end{aligned}
$$ {#eq-log-logistic-dens}

[^scaling]: If $T$ has log-logistic distribution with density $f_0$, then $e^{\eta} T$ is log-logistic distributed with density $f_{\eta}$, cf. @exr-loglog. 
::: 

\index{accelerated failure time} 

For maximum likelihood estimation 
we need to compute and optimize the likelihood in @eq-survlike. Efficient 
computation requires access to the density as well as the survival function, 
which are both explicitly available for the log-logistic distribution, 
which makes this distribution attractive for AFT models. 

::: {#exm-log-logistic-prostate}

We implement the negative log-likelihood (`nll()`) for the AFT model with 
a log-logistic survival distribution below. The implementation is 
based on @eq-survlike using the expressions ([-@eq-log-logistic-surv])
and ([-@eq-log-logistic-dens]) for the survival function and density, 
respectively, of the log-logistic distribution. 

The `nll()` function takes four arguments. Besides a vector of parameters (`par`),
the censored survival times (`time`) and the status indicator (`status`), it 
takes a single additional argument `x` representing a categorical variable 
with $k$ values. The parameter vector `par` is of the form 
$$
    (\sigma, \beta_0, \ldots, \beta_{k-1}) = (\lambda^{-1}, \beta_0, \ldots, \beta_{k-1})
$$
with $\sigma$ the scale parameter, $\beta_0$ the intercept and $\beta_1, \ldots, \beta_{k-1}$ 
are contrasts. The survival function is also implemented below for later usage.

```{r}
#| label: log-logistic-log-likelihood
#| code-fold: show
# Argument `x` is an integer representing the k levels of a single 
# categorical predictor variable. `par[2:(k+1)]` are the k 
# beta-parameters corresponding to the intercept and contrasts. 
# Its default value corresponds to no predictor.
nll <- function(par, time, status, x = rep(1, length(time))) {
    lambda <- par[1]^{-1}
    eta <- par[2] + c(0, par[-c(1, 2)])[x] # beta_x
    uncens <- status == 1
    lt <- log(time)
    logf <- (log(lambda) - lambda * eta[uncens] + (lambda - 1) * lt[uncens] - 
        2 * log(1 + exp(lambda * (lt[uncens] - eta[uncens])))) |> sum()
    logS <- log(1 + exp(lambda * (lt[!uncens] - eta[!uncens]))) |> sum()
    - logf + logS
}
```

We use `optim()` with method `L-BFGS-B` to compute the maximum-likelihood estimate
of a single survival distribution. Since some of the patients have survived less 
than one months, some survival times are registered as $0$, which is first of 
all numerically problematic (due to $\log(0)$). For this reason we offset all times by 
$0.5$ corresponding to shifting all death times by half a months.

```{r}
#| label: log-logistic-optim
#| code-fold: show
prostate_aft <- optim(
    c(2, 2), 
    nll, 
    method = "L-BFGS-B", 
    lower = c(0.01, -Inf),
    time = prostate$dtime + 0.5, # to avoid log(0)
    status = prostate$status != "alive"
)
prostate_aft[1:2]
```

```{r}
#| label: fig-log-logistic-prostate
#| fig-cap: "The survival function for the fitted log-logistic distribution (blue) and the Kaplan-Meier estimator (black)."
#| column: margin
#| echo: false
#| opts.label: smallfigure
S <- function(t, sigma, eta) 1 / (1 + exp((log(t) - eta) / sigma))
fit <- tibble(time = 0:80, surv = S(time + 0.5, prostate_aft$par[1], prostate_aft$par[2]))
summary(prostate_KM, data.frame = TRUE) |> 
        ggplot(aes(time + 0.5, surv)) + 
        geom_step() +
        geom_line(data = fit, color = "blue") + 
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability")

```


We see from the `optim()` result that the parameter estimates are $\hat{\sigma} =$ `r round(prostate_aft$par[1], 2)`
and $\hat{\beta}_0 =$  `r round(prostate_aft$par[2], 2)`, and 
we can also see that $-\ell(\hat{\sigma}, \hat{\beta}_0) =$ `r round(prostate_aft$value, 2)`
The corresponding fitted survival function is compared to the Kaplan-Meier 
estimator in @fig-log-logistic-prostate.

We can also use our implementation to fit the regression model where the 
survival distribution depends on treatment.

```{r}
#| label: log-logistic-optim-rx
#| code-fold: show
prostate_aft_rx <- optim(
    c(2, 2, 0, 0, 0), 
    nll, 
    method = "L-BFGS-B", 
    lower = c(0.1, -Inf, -Inf, -Inf, -Inf),
    time = prostate$dtime + 0.5, # to avoid log(0)
    status = prostate$status != "alive", 
    x = as.numeric(prostate$rx)
)
prostate_aft_rx[1:2]
```



```{r}
#| label: fig-prostate-aft-rx
#| fig-cap: "The AFT survival functions for the fitted log-logistic distributions (full) and the Kaplan-Meier estimators (dashed) for each of the four treatment groups."
#| fig-cap-location: margin
# The log-logistic survival function in the scale-location parametrization. 
time0 <- 0:80
fit <- tibble(
    time = rep(time0, 4), 
    surv = c(
        S(time0 + 0.5, prostate_aft_rx$par[1], prostate_aft_rx$par[2]),
        S(time0 + 0.5, prostate_aft_rx$par[1], prostate_aft_rx$par[2] + prostate_aft_rx$par[3]),
        S(time0 + 0.5, prostate_aft_rx$par[1], prostate_aft_rx$par[2] + prostate_aft_rx$par[4]),
        S(time0 + 0.5, prostate_aft_rx$par[1], prostate_aft_rx$par[2] + prostate_aft_rx$par[5])
    ),
    strata = rep(paste("rx=", levels(prostate$rx), sep = ""), each = 81)
)

summary(prostate_KM_rx, data.frame = TRUE) |> 
        ggplot(aes(time + 0.5, surv, color = strata)) + 
        geom_step(linetype = 2) +
        geom_line(data = fit) + 
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top")
```
:::

We can fit AFT models, and a range of other parametric regression models, via the 
`survreg()` function from the survival package. 

```{r}
#| label: prostate-survreg-rx
#| echo: true
survreg(
    Surv(dtime + 0.5, status != "alive") ~ rx, 
    data = prostate, 
    dist = "loglogistic"
)
```

Note that the parametrization used by `survreg()` is exactly the same as we implemented, 
and that the estimated parameter values and computed log-likelihood are the same as well
(up to small numerical differences). 

The `survreg()` function even gives a likelihood-ratio test statistic (`Chisq`) of the hypothesis 
that there are no differences between the treatments. The corresponding 
$p$-value is computed using an approximating $\chi^2$-distribution with 
$3$ degrees of freedom. It is above $0.1$ and we cannot reject that 
there is no treatment effect. 

As for `lm` and `glm` objects there are methods for `survreg` objects implemented 
in the broom package fo the `glance()` and `tidy()` functions and in the survival 
package for `anova()`. We will illustrate some of their usages in subsequent examples.

#### Proportional hazards models {.unnumbered}

A different model class is given by letting the linear predictor have
a multiplicative effect on the hazard function. The resulting 
proportional hazards models are among the most widely 
used survival models. 

\index{proportional hazards model} \index{baseline hazard}

::: {#def-prop-hazard} 
With $\lambda_0$ a baseline hazard function, the proportional hazards model 
is given by the hazard function
$$
    \lambda(t) = \lambda_0(t) e^{\eta}
$$
with $\eta$ the linear predictor. 
:::

It follows that for the cumulative hazard function, 
$$
    \Lambda(t) = \Lambda_0(t) e^{\eta},
$$
and the proportionality holds for the cumulative hazard as well. 
The factor $e^{\beta_j}$ is the *hazard ratio* \index{hazard ratio} between two models
corresponding to a unit change of the $j$-th predictor.

The proportional hazards model is sometimes called the *Cox model*. 
@Cox:1972 discovered that for the proportional hazards model, 
the parameters that enter into the linear predictor can be estimated 
without specifying the baseline hazard function. This will be covered in @sec-semipar-surv. 
Here we will instead show how the parameters 
can be estimated via maximum-likelihood when the baseline is assumed to 
be a Weibull distribution. 

::: {#exm-weibull-prop}
The Weibull baseline hazard function and cumulative hazard function are  
$$
    \lambda_0(t) = \gamma t^{\gamma-1} \quad  \text{and} \quad \Lambda_0(t) = t^{\gamma}.
$$
The log-likelihood is 
$$
    \begin{align*}
    \ell & = \sum_{i=1}^n \delta_i \log(\gamma T_i^{\gamma-1}e^{\eta_i}) - T_i^{\gamma}e^{\eta_i} \\ 
    & = \underbrace{\sum_{i=1}^n \delta_i \log(T_i^{\gamma}e^{\eta_i}) - T_i^{\gamma}e^{\eta_i}}_{\text{Poisson log-likelihood}} + 
    \sum_{i=1}^n \delta_i \log(\gamma T_i^{-1}).
    \end{align*}
$$
This is (surprisingly) up to a constant the log-likelihood for a
Poisson model of the $\delta_i$-s with logarithmic link function and mean value 
$T_i^{\gamma}e^{\eta_i}$ for fixed $\gamma$. Note that the $\alpha$ parameter 
appearing in @exm-weibull-hazard is captured by an intercept in the linear predictor.

For a fixed value of the linear predictor we also find that 
$$
    \partial_{\gamma} \ell = \sum_{i=1}^n (\delta_i- T_i^{\gamma} e^{\eta_i}) \log T_i + \frac{\delta_i}{\gamma}.
$$
Thus $\gamma$ solves the equation
$$
    \gamma = \frac{n_u}{\sum_{i=1}^n (T_i^{\gamma} e^{\eta_i} -\delta_i)\log T_i}
$$ {#eq-gamma-mle}
with $n_u$ the number of uncensored observations.
[A practical algorithm can alternate between: i) estimate the linear predictor 
as $\hat{\eta}^{(k)}$ by optimizing the Poisson log-likelihood (for fixed 
$\gamma^{(k-1)}$); and ii) update
$$
    \gamma^{(k)} = \frac{n_u}{\sum_{i=1}^n (T_i^{\gamma^{(k-1)}} e^{\hat{\eta}_i^{(k)}} -\delta_i)\log T_i}.
$$
]{.aside}
::: 

The example above shows how implementations of generalized linear models -- 
and Poisson regression models, in particular -- 
can be used to fit the Weibull proportional hazards model (for fixed $\gamma$)
with the survival times entering as an offset term. This was of historical 
importance as it made computations practical early on using some standard 
algorithms available. Today, the connection between the survival likelihood 
for Weibull distributions and the Poisson likelihood
is mostly of theoretical and historical interest.

To fit the model we will instead show that the proportional hazards model with a Weibull baseline 
hazard function is also an accelerated failure time model, which makes it possible to 
use `survreg()` to fit the model. 

::: {#exm-weibull-aft}

If we take $G$ in @def-AFT to be the extreme value distribution 
$$
    G(z) = 1 - \exp(-e^{z}),
$$ 
then the corresponding AFT model has survival function 
$$
    S_\eta(t) = \exp\left(-(e^{-\eta}t)^{\frac{1}{\sigma}}\right) = \exp\left(-e^{-\gamma \eta}t^\gamma\right)
$$
where $\gamma = \frac{1}{\sigma}$. That is, the cumulative hazard function is 
, which is the proportional hazards 
model with a Weibull baseline, but reparametrized in terms of the scale 
parameter $\sigma = \frac{1}{\gamma}$, and the linear predictor is 
multiplied by $-\gamma$. 

We can use the `survreg()` function to fit the model. 

```{r}
#| label: tbl-prostate-weibull
#| tbl-cap: "Parameter estimates for the Weibull accelerated failure time model fitted to the prostate cancer survival data."
#| tbl-cap-location: bottom
#| code-fold: show
prostate_survreg <- survreg(
    Surv(dtime + 0.5, status != "alive") ~ rx, 
    data = prostate, 
    dist = "weibull"
)
prostate_survreg |> tidy()
```

We can compute the estimated survival functions for the Weibull model 
much in the same way as we did above for the log-logistic model, where 
we used the MLE of the parameters and the explicit formula for the 
survival function. However, this is a bit cumbersome and does not generalize 
well. We can instead use the `predict()` method for `survreg` objects 
to predict quantiles for the survival functions, which implicitly also 
gives the estimated survival functions.

```{r}
#| echo: false
#| eval: false
S <- function(t, sigma, eta) exp( - exp((log(t) - eta) / sigma))
beta <- coef(prostate_survreg)

time0 <- 0:80
fit <- tibble(
    time = rep(time0, 4), 
    surv = c(
        S(time0 + 0.5, prostate_survreg$scale, beta[1]),
        S(time0 + 0.5, prostate_survreg$scale, beta[1] + beta[2]),
        S(time0 + 0.5, prostate_survreg$scale, beta[1] + beta[3]),
        S(time0 + 0.5, prostate_survreg$scale, beta[1] + beta[4])
    ),
    strata = rep(paste("rx=", levels(prostate$rx), sep = ""), each = 81)
)

summary(prostate_KM_rx, data.frame = TRUE) |> 
        ggplot(aes(time + 0.5, surv, color = strata)) + 
        geom_step(linetype = 2) +
        geom_line(data = fit) + 
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top") +
        xlim(0, 82)
```


```{r}
#| label: fig-prostate-weibull-rx
#| fig-cap: "The survival functions for the fitted Weibull distributions (full) and the Kaplan-Meier estimators (dashed) for each of the four treatment groups."
#| fig-cap-location: margin
fit <- tibble(rx = factor(1:4, labels = levels(prostate$rx)))
p_seq <- seq(0.01, 0.90, 0.01)
pred <- predict(prostate_survreg, newdata = fit, type = "quantile", p = p_seq)
fit <- cbind(fit, as_tibble(pred, .name_repair = "minimal")) |> 
    pivot_longer(cols = !rx, values_to = "time") |>
    mutate(
        surv = 1 - rep(p_seq, 4),
        strata = paste("rx=", rx, sep = "")
    ) |> 
    filter(time <= 80)

summary(prostate_KM_rx, data.frame = TRUE) |> 
        ggplot(aes(time + 0.5, surv, color = strata)) + 
        geom_step(linetype = 2) +
        geom_line(data = fit, aes(time)) + 
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top") +
        xlim(0, 82)
```

@fig-prostate-weibull-rx compares the fitted Weibull distributions to the 
Kaplan-Meier estimators for the four treatment groups, and @tbl-prostate-weibull-test 
reports the test statistic and $p$-value for testing if there are no 
differences between the four treatment groups. The $p$-value is a bit smaller 
in this case than when using the log-logistic distribution, and this time 
just below 5%. 

```{r}
#| label: tbl-prostate-weibull-test
#| tbl-cap: "Deviance (likelihood-ratio) test of the hypothesis that there are no difference between the three treatment groups."
#| tbl-cap-location: bottom
#| column: margin
prostate_survreg |> 
    anova() |> 
    select(-4) |> 
    tidy() |> 
    filter(term == "rx") |> 
    knitr::kable(digits = 3)
``` 

To corroborate the technical relation between the log-linear Poisson regression model and 
the proportional hazards model with a Weibull baseline, we refit the $\beta$-parameters
using `glm()` (fixing the scale parameter at its estimated value).

```{r}
#| label: tbl-prostate-glm
#| tbl-cap: "Regression parameters estimated using `glm()` and identical to those in @tbl-prostate-weibull."
#| tbl-cap-location: bottom
#| code-fold: show
sigma <- prostate_survreg$scale
glm((status != "alive") ~ offset(log(dtime + 0.5) / sigma) + rx, 
    data = prostate, 
    family = "poisson"
) |> coefficients() * (- sigma)
```
:::

To conclude our analyses of the prostate survival times, we have shown how one 
treatment group (1.0 mg estrogen) has a slightly different survival distribution 
than the other three with a better survival prognosis. The statistical tests 
of any treatment effect, based on two parametric models, are at best borderline significant
though, and we cannot draw any clear conclusion. 

There do not seem to be any detectable difference between the other three treatment groups. 
It is particularly noteworthy that the treatment group with the highest dose (5.0 mg estrogen)
have a survival distribution simular to the placebo group. If there were an effect, we 
would often expect it to be monotone in dosage, and it would be peculiar if the effect is 
only there for the medium dose. 

## Semiparametric survival models {#sec-semipar-surv}

For the parametric proportional hazards models in @def-prop-hazard, the baseline 
hazard function $\lambda_0$ is either known or parametrized itself. There is a risk 
that we have misspecified this baseline model, and we would like to avoid 
a model of the baseline hazard if possible. This leads to the semiparametric proportional 
hazards model, where an individual's hazard is proportional to a 
common baseline hazard $\lambda_0$ with an individual weight factor, but where we 
do not model[^semi] the baseline $\lambda_0$. 

[^semi]: This is similar to *not* making distributional assumptions such as A3 or GA3 for the linear and generalized models.

The general proportional hazards model assumes that the 
dependence of the predictors is via a multiplicative factor on the baseline, 
and the $i$-th individual thus has hazard function
$$
    \lambda_i(t) = w_i \lambda_0(t)
$$
for $w_i > 0$ a weight[^weight]. The cumulative hazard function is consequently $\Lambda_i(t) = w_i
\Lambda_0(t)$.

[^weight]: The $i$-th weight could be $e^{\eta_i}$ with $$\eta_i = X_i^T \beta$$ the linear predictor, but it does not need to be for the general arguments.

::: {#def-cum-weight} 
In terms of the observations $T_1, \ldots, T_n$ and weights $w_1, \ldots, w_n \geq 0$ we 
define the cumulative weight process 
$$
    W(t) = \sum_{j: t \leq T_j} w_j
$$ 
and the cumulative weights $W_i = W(T_i)$.
:::

::: {#prp-fac2}
When $\Lambda_i(t) = w_i \Lambda_0(t)$ the likelihood equals
$$
    L = \prod_{i: \delta_i = 1} \frac{w_i}{W_i} \left(\prod_{i} 
    (W_i \lambda_0(T_i))^{\delta_i}\right) e^{-\int_0^{\infty} W(t) \lambda_0(t) \mathrm{d} t}.
$$
[To ease notation we leave out limits on the generic index $i$ in products and sums. It will always run through observations from $1$ to $n$.]{.aside}
::: 

::: {.proof}
First note that with $W(t) = \sum_{j: t \leq T_j} w_j$,
$$
    \sum_{i} \Lambda_i(T_i) = \int_0^{\infty} W(t) \lambda_0(t) \mathrm{d} t.
$$
The likelihood can therefore be factorized as  
$$
    \begin{align*}
      L & = \prod_{i} \lambda_i(T_i)^{\delta_i} e^{-\Lambda_i(T_i)} \\
        & = \left(\prod_{i: \delta_i = 1} (w_i \lambda_0(T_i))^{\delta_i}\right) e^{-\int_0^{\infty} W(t) \lambda_0(t)
        \mathrm{d} t} \\ 
        & = \prod_{i: \delta_i = 1} \frac{w_i}{W_i} \left(\prod_{i} 
        (W_i \lambda_0(T_i))^{\delta_i}\right) e^{-\int_0^{\infty} W(t) \lambda_0(t)
        \mathrm{d} t}.
    \end{align*}
$$
:::

The purpose of @prp-fac2 is to decompose the likelihood into two factors, 
where the second factor
$$
    \tilde{L}(\lambda_0) = 
    \left(\prod_{i} (W_i \lambda_0(T_i))^{\delta_i}\right) e^{-\int_0^{\infty} W(t) \lambda_0(t) \mathrm{d} t}.
$$
is the only one depending on $\lambda_0$. This factor can be made arbitrarily large if 
we try to maximize over $\lambda_0$, and there is thus no unconstrained 
"maximum-likelihood" estimator of $\lambda_0$ in a classical sense.

It is possible to formalize, via a certain renormalization, that for fixed weights 
the "maximizer" is a generalized function of the form 
$$
    \hat{\lambda}_0 = \sum_{i} \frac{\delta_i}{W_i} \delta_{T_i}
$$
where $\delta_{T_i}$ is the Dirac measure in $T_i$. Since
$$
\int_0^{\infty} W(t) \hat{\lambda}_0 \mathrm{d} t = \sum_{i} W_i \frac{\delta_i}{W_i} = \sum_{i} \delta_i = n_u,
$$
the number of uncensored obervations, this leads to 
$$
    \tilde{L}(\hat{\lambda}_0) = e^{-n_u},
$$
which is independent of the weights. We pursue a similar perspective in @sec-pois-emp-like.


### Cox's partial likelihood

@prp-fac2 and the subsequent discussion suggest that for estimation 
of the parameters that enter into the weights[^var-ind2], and not into 
$\lambda_0$, we can ignore the factor $\tilde{L}$ in the likelihood.
This motivates our definition of Cox's *partial* likelihood. 

[^var-ind2]: And are variation independent with $\lambda_0$.

\index{Cox's partial likelihood} \index{partial likelihood}

::: {#def-cox-likelihood}
Cox's partial likelihood for the proportional hazards model is 
$$
    L_{\mathrm{par}} = \prod_{i: \delta_i = 1} \frac{w_i}{W_i}
$$
where the cumulative weights $W_i$ are given by @def-cum-weight.
::: 

In practice, the predictors are mostly linked to the weights via an exponential 
function and are thus of the form
$$
    w_i = e^{X^T_i \beta} = e^{\eta_i}.
$$ {#eq-exp-weights}
In this case we find that the partial log-likelihood equals   
$$
    \ell_{\mathrm{par}}(\beta) = 
    \sum_{i: \delta_i = 1} X_i^T \beta - \log \Bigg(\underbrace{\sum_{j: T_i \leq T_j} w_j}_{W_i} \Bigg),
$$ {#eq-partial-ll}
with corresponding score function 
$$
    \mathcal{U}_{\mathrm{par}}(\beta) = D \ell_{\mathrm{par}}(\beta) = \sum_{i: \delta_i = 1} X_i^T - 
    \frac{\sum_{j: T_i \leq T_j} X_j^T w_j}{W_i}
$$ {#eq-partial-score}
and information function

$$
    \begin{aligned}
       \hskip -30mm \mathcal{J}_{\mathrm{par}}(\beta) & = - D^2  \ell_{\mathrm{par}}(\beta) \\
        & = \sum_{i: \delta_i = 1} \frac{\sum_{j: T_i \leq T_j} X_j X_j^T w_j}{W_i} - 
        \frac{(\sum_{j: T_i \leq T_j} X_j w_j) (\sum_{j: T_i \leq T_j} X_j w_j)^T}{W_i^2}.
    \end{aligned}
$$ {#eq-partial-information}

The estimate, $\hat{\beta}$, is computed by maximizing $\ell_{\mathrm{par}}$, and 
the observed information, $\mathcal{J}(\hat{\beta}),$ is used for quadratic approximations 
of $\ell_{\mathrm{par}}$, for standard error estimates and Wald tests. In practice, the computations 
proceed just as if $\ell_{\mathrm{par}}$ were a log-likelihood function. 

The particular score test
$$
    \mathcal{U}(0)^T \mathcal{J}(0)^{-1} \mathcal{U}(0)
$$
for testing $\beta = 0$ reduces to a classical test known as the *log-rank test* \index{log-rank test} 
in the special case with only a single categorical predictor.  

::: {#exm-prostate-coxph}
Cox's proportional hazards model can be fitted to data via the function `coxph()` 
from the survival R package. 

```{r}
#| label: tbl-prostate-cox
#| tbl-cap: "Estimated regression parameters for Cox's proportional hazards model for the prostate cancer survival data."
#| tbl-cap-location: bottom
#| code-fold: show
prostate_cox <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ rx, 
    data = prostate, 
)
prostate_cox |> tidy()
```


```{r}
#| label: tbl-prostate-cox-test
#| tbl-cap: "Likelihood-ratio test of the hypothesis of no treatment effect."
#| tbl-cap-location: bottom
#| column: margin
anova(prostate_cox) |> 
    tidy() |> 
    filter(term == "rx") |> 
    knitr::kable(digits = 3)
```


```{r}
#| label: cox-survdiff-comp
#| eval: false
#| echo: false
survdiff(
    Surv(dtime, status != "alive") ~ rx, 
    data = prostate,
)
summary(prostate_cox)
```

The parameter estimates are comparable to those obtained by `survreg()` with the 
Weibull distribution, by appropriately rescaling, since that model is parametrized 
as an AFT model. In particular, they Cox model regression parameters have the opposite 
sign. This is noteworthy, since a negative value of a Cox model regression parameter 
means that the corresponding predictor has a positive association with the survival time. 
That is, if $\beta_i < 0$ then a unit increase of $X_i$ decreases the hazard by a factor 
$e^{-\beta_i} < 1$, which results in a survival distribution with longer survival times. 

Note that the (partial) likelihood-ratio test reported in @tbl-prostate-cox-test results in a very 
similar conclusion as using the Weibull baseline.
:::

### Poisson empirical likelihood {#sec-pois-emp-like}

To provide an additional perspective on Cox's partial likelihood, 
we will show how it can also be derived from a rigorous profiling 
argument of the Poisson empirical likelihood we define below. 
Our main purpose of giving this argument is that we then also get 
an explicit formula for a nonparametric estimator of the
cumulative baseline hazard function. 

For $\lambda_1, \ldots, \lambda_n \geq 0$ we define the corresponding *jump cumulative hazard function* as 
$$
    \Lambda(t) = \sum_{i : T_i \leq t} \lambda_i.
$$
We can regard this cumulative hazard function as having the hazard function
$$
    \lambda(t) = \sum_{i} \lambda_i \delta_{T_i},
$$
which is, of course, not a function but a measure, and $\Lambda(t)$ is the 
measure of the interval $[0, t]$. The measure has point mass $\lambda_i$ 
in the jump point $T_i$. 

::: {#def-pois-surv-likelihood}
The Poisson empirical likelihood of a jump cumulative hazard function is 
$$
    L^*(\Lambda) = \prod_{i}  \lambda_i^{\delta_i} e^{- \Lambda_i}
$$
where $\Lambda_i = \Lambda(T_i)$.
:::

In this section we think about $L^*$ as simply a formal extension of the likelihood ([-@eq-survlike2])
from continuously differentiable cumulative hazard functions to jump functions, 
but see also @sec-dis-surv on discrete time survival distributions, and see
@Johansen:1983 for a justification of $L^*$ as a real likehood.

The next lemma gives an expression for the Poisson empirical log-likelihood 
for jump cumulative hazard functions of the form
$$
    \Lambda_i(t) = w_i \sum_{j: T_j \leq t} e^{h_j}.
$$

::: {#lem-poisson-prop} 
Suppose that $\Lambda_0$ has jumps $\lambda_i = e^{h_i}$ for $h_i \in \mathbb{R}$. 
For the proportional hazards model with baseline $\Lambda_0$ the Poisson empirical log-likelihood is
$$
    \ell^* = \sum_{i} \delta_i h_i + \delta_i \log (w_i) - e^{h_i} W_i.
$$ {#eq-ell-star}
:::

::: {.proof}
The Poisson log-likelihood for the proportional hazards model is 
$$
    \begin{align*}
        \ell^* & = \sum_{i} \delta_i \log(\lambda_0(T_i)) + \delta_i \log (w_i) - \Lambda_0(T_i)w_i \\
        & = \sum_{i} \delta_i h_i + \delta_i \log(w_i) - w_i \sum_{j : T_j \leq T_i} e^{h_j},
    \end{align*}
$$ 
where $\ell^*$ is expressed in terms of the weights and the exponents $h_j$. Interchanging the sums in the log-likelihood we get 
$$
    \ell^* = \sum_{i} \delta_i h_i + \delta_i \log (w_i) - e^{h_i} W_i.
$$ 
:::

::: {#lem-profile} 
The maximizer of $\ell^*$ in ([-@eq-ell-star]) over $h_1, \ldots, h_n$ is given by 
$$
    e^{h_i} = \frac{\delta_i}{W_i}
$$
for $i = 1, \ldots, n$. 
[With all weights being $1$ in @lem-profile, the estimator is recognized as the Nelson-Aalen estimator.]{.aside}
:::

::: {.proof}
The parameters $h_i$ are variation independent, and we
  can thus maximize the function by maximizing each term. Differentiating we find that 
$$
    \partial_{h_i} \ell^* = \delta_i - e^{h_i} W_i.
$$
If $\delta_i = 0$ the function is monotonely decreasing and maximized 
in $- \infty$ corresponding to $e^{h_i} = 0$. If $\delta_i = 1$ we equate 
the derivative equal to $0$ and find the claimed solution to be the
unique stationary point. An additional
differentiation shows that the second derivative is 
$$
    - e^{h_i} W_i  < 0, 
$$
and this implies that the solution is a global maximum. 
::: 

As an immediate consequence of plugging the maximizer into $\ell^*$ 
we get the following corollary.

::: {#cor-emp-profile}
The profile log-likelihood for the weights of the Poisson empirical log-likelihood is 
$$
    \ell^*_{\mathrm{profile}} = \underbrace{\sum_{i: \delta_i = 1} \log \left(
    \frac{w_i}{W_i}\right)}_{\text{\hskip -5mm Cox's partial log-likelihood}} - \ n_u
$$ 
where $n_u = \sum_{i} \delta_i$ is the total number of uncensored observations 
(the number of deaths).
:::

Corollary @cor-emp-profile shows how Cox's partial likelihood 
appears as a profile likelihood function from the Poisson empirical likelihood 
for the proportional hazards model. 

When $w_i = e^{X_j^T \beta}$, @lem-profile results in the estimator 
$$
    \hat{\Lambda}_0(t) = \sum_{i: T_i \leq t} \frac{\delta_i}{\sum_{j:T_i \leq T_j} e^{X_j^T \hat{\beta}}}
$$
of the cumulative hazard function in terms of the survival times and the 
estimate $\hat{\beta}$. This estimator is known as the *Breslow estimator*,
which can be seen as an extension of the Nelson-Aalen estimator of the baseline 
cumulative hazard function. 

\index{Breslow estimator} 



::: {#exm-prostate-breslow}

We continue @exm-prostate-coxph and compute the Breslow estimator of the baseline 
cumulative hazard using `survfit()`, see `?survfit.coxph` for details. 
In fact, `survfit()` returns an estimate of the baseline survival function as
$$
    \hat{S}_0(t) = e^{-\hat{\Lambda}_0}
$$
where $\hat{\Lambda}_0$ is the Breslow estimator.


```{r}
#| label: fig-prostate-cox-breslow
#| fig-cap: "The Breslow estimate of the baseline survival function."
#| column: margin
#| code-fold: show
survfit(prostate_cox) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time + 0.5, surv)) + 
        geom_ribbon(aes(ymin = lower, ymax = upper), fill = "gray90") +
        geom_step() +
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") 
```

@fig-prostate-cox-breslow shows the Breslow estimate including a pointwise 
95% confidence band, which is also computed by the `survfit()` function. 
@fig-prostate-coxph-rx shows the fitted survival functions for the four different
treatment groups, using the Breslow estimator of the baseline cumulative hazard function,
compared to using a Weibull parametric model as baseline. The survival curves are 
based on the Cox model and in principle computed as 
$$
    \hat{S}_k = e^{-e^{\hat{\beta}_k} \hat{\Lambda}_0}, \quad k = 0, 1, 2, 3.
$$ {#eq-mult-baseline}
Note that for Cox's proportional hazards model, $\beta_0 = 0$ always. That is, there is no 
intercept parameter. Since the baseline hazard is nonparametric, we cannot identify a separate 
intercept parameter in this model, and the intercept is effectively absorbed into the nonparametric 
baseline. As pointed out in the  `survfit()` documentation, this also means that there is no 
such thing as *the* baseline hazard. 
What the baseline hazard is depends on a choice of baseline value of $X$, and thus also 
on how the predictors are encoded. In this example, the 
baseline computed by `survfit()` corresponds to the placebo treatment group, but in more complicated setups the baseline 
corresponds to a pseudo individual -- see `?survfit.coxph`. 

While it can be instructive to extract the estimated regression parameters and use
([-@eq-mult-baseline]) explicitly, the code below used for actually computing the 
estimated survival functions shown in @fig-prostate-coxph-rx use `survfit()` with 
a new dataset, which will give a survival function for each row of the new data. 
This is both a safer and more convenient way of computing such estimates. 

```{r}
#| echo: false
#| eval: false
beta <- coefficients(prostate_cox)
summary(prostate_cox_baseline, data.frame = TRUE) |> 
    mutate(
        cumhaz1 = exp(beta[1]) * cumhaz,
        cumhaz2 = exp(beta[2]) * cumhaz,
        cumhaz3 = exp(beta[3]) * cumhaz
    ) |> 
    pivot_longer(
        cols = c(cumhaz, cumhaz1, cumhaz2, cumhaz3), 
        values_to = "cumhaz", 
        names_to = "strata"
    ) |> 
    mutate(
        strata = factor(strata, labels = paste("rx=", levels(prostate$rx), sep = ""))
    ) |>
    mutate(surv = exp(-cumhaz)) |>
    ggplot(aes(time + 0.5, surv, color = strata)) + 
        geom_line(data = fit, aes(time), linetype = 2) +
        geom_step() +
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top")
```
:::


```{r}
#| label: fig-prostate-coxph-rx
#| fig-cap: "The fitted survival functions for proportional hazards models using the Cox proportional hazards model with the Breslow estimator of the baseline (full, step) and the parametric Weibull survival functions (dashed) for each of the four treatment groups."
#| fig-cap-location: margin
#| message: false
pred_frame <- tibble(
    data = 1:4,
    rx = factor(1:4, labels = levels(prostate$rx)),
    strata = factor(1:4, labels = paste("rx=", rx, sep = ""))
    )
survfit(prostate_cox, newdata = pred_frame) |>
    summary(data.frame = TRUE) |>
    left_join(pred_frame) |>
     ggplot(aes(time, surv, color = strata)) + 
        geom_line(data = fit, linetype = 2) +
        geom_step() +
        ylim(0, 1) +
        xlab("time (months)") + 
        ylab("survival probability") +
        theme(legend.position = "top")
```


### Ties

We did not deal with ties explicitly in any of the derivations in previous sections. 
If there are ties among the survival times, the summation over $j: T_i \leq T_j$ in the 
definition of the cumulative weights is still well defined. If it happens that $T_i = T_k$ 
the corresponding cumulative weights $W_i = W_k$ also become identical.
And if both observations are uncensored, they enter into the partial likelihood 
as the product
$$
    \frac{w_i}{W_i} \frac{w_k}{W_k} = \frac{w_i w_k}{W_i^2}.
$$
This implicit way of handling ties is known as Breslow's method. It 
treats ties as if they were really due to events that happend at the exact same time.

If the tie $T_i = T_k$ was due to rounding or grouping, the contribution 
to the partial likelihood from those two observations should have been one of the two factors:
$$
    \frac{w_i w_k}{W_i (W_i - w_i)} \quad \text{or} \quad \frac{w_i w_k}{W_k (W_k - w_k)}.
$$ {#eq-fac}

We could try to break the tie, e.g., by letting the R function `order()` break ties in whatever 
arbitrary way it chooses depending on the sorting algorithm and the original order of the data.
This naive solution will break all ties but in a unsystematic way, choosing one of the two factors above 
arbitrarily. Breaking ties in this way appears difficult to justify, and we should probably 
only break ties if we actually know and use the correct ordering prior to the rounding. 

\index{ties!Efron's method}

Efron's method[^default] represents instead the "correct but unknown" factor
by an approximation. Efron's approximation for two ties is 
$$
    \frac{w_i w_k}{W_i (W_i - (w_i + w_k)/2)} = \frac{w_i w_k}{W_k (W_k - (w_i + w_k)/2)}
$$
with similar formulas for more than two ties. The approximation is 
close to the geometric mean of the factors ([-@eq-fac]).

[^default]: The default in `coxph()` for handling ties is Efron's method.

Results rarely depend much on whether Efron's or Breslow's method (or even naive tie breaking) 
is used, but Efron's method is usually preferred if the ties are due to a lack of precision 
and not to a true discrete nature of the survival times. 


## Survival residuals {#sec-res-surv}

Model fit can be assessed for survival regression models just as for generalized 
linear models by means of residuals. There are, of course, several different possible choices. 

### Cox-Snell residuals

Cox-Snell residuals are not really residuals in a classical sense, but a transformation 
of the observed survival times to a set of random variables whose distribution should be 
exponential if the model is correct. They are based on the following result.

::: {#lem-cum-haz-trans}
If $T$ has continuous cumulative hazards function $\Lambda$ 
then  $\Lambda(T)$ is exponentially distributed with rate 1. 
::: 

::: {.proof}
The survival function $S(t) = \exp(-\Lambda(t))$ is continuous, whence  
$$
    S(T) \sim \text{unif}([0,1]).
$$
From this it follows that 
$$
    \mathbf{P}(\Lambda(T) \leq t) = \mathbf{P}(S(t) > \exp(-t)) = 1 - \exp(-t)
$$
for $t \geq 0$, or $\Lambda(T) \sim \text{Exp}(1)$.
:::

::: {#def-CS}
With $\hat{\Lambda}_i$ an estimator of the cumulative hazard function for $T_i$ the
Cox-Snell residuals are defined as \index{residual!Cox-Snell}
$$
    \hat{\Lambda}_i(T_i)
$$ {#eq-CS}
for $i = 1, \ldots, n$. 
::: 

From the lemma above we conclude that if $(T_1, \delta_1), \ldots, (T_n, \delta_n)$ are independently right censored
 survival times  with corresponding hazard functions $\Lambda_1, \ldots, \Lambda_n$, then 
$$
    (\Lambda_1(T_1), \delta_1), \ldots, (\Lambda_n(T_n), \delta_n)
$$
are independently right censored i.i.d. exponentially distributed variables. 

If there were no censoring, we would expect the Cox-Snell residuals to follow approximately an
exponential distribution. With censoring the Cox-Snell residuals are expected to 
be approximately independently censored exponentially distributed. We use this 
in practice by estimating the cumulative hazards function for the Cox-Snell residuals 
using, e.g., the Nelson-Aalen estimator or the negative log of the Kaplan-Meier 
estimator. The resulting estimate is then 
compared to the cumulative hazard for the exponential distribution, which
is $\Lambda(t) = t$. In practice we plot the estimated cumulative 
hazards against the Cox-Snell residuals, which should fall on 
a straight line with slope $1$. 

### Martingale residuals 

The martingale residuals are defined as \index{residual!martingale}
$$
    \delta_i - \hat{\Lambda}_i(T_i),
$$
and are thus the status indicators minus the Cox-Snell residuals. 
Their usefulness is partly justified by the following result.

::: {#lem-MG}
If $T = \min\{T^*, C\}$ with $T^*$ and $C$ having continuously
differentiable survival functions it holds that
$$
    \mathbf{E}(\Lambda(T)) = \mathbf{P}(T^* \leq C).
$$ 
:::

::: {.proof}
Let $S$ and $H$ denote the survival functions of $T^*$ and $C$, 
respectively. By integration by parts
$$
    \begin{align*}
        \mathbf{E}(\Lambda(T)) & = - \int_0^{\infty} \Lambda(t) (SH)'(t) \mathrm{d} t =
        \int_0^{\infty} \Lambda'(t) S(t) H(t) \mathrm{d} t \\
        &  = \int_0^{\infty} f(t) H(t) \mathrm{d} t = \mathbf{P}(T^* \leq C).
    \end{align*}
$$
[The survival function for the minimum of the two independent random 
variables $T^*$ and $C$ is the product $SH$, whence the density of their distribution 
is the negative derivative $-(SH)'$.]{.aside}
:::

It follows from @lem-MG that with $\delta = 1(T^* \leq C)$ then 
$$
    \mathbf{E} (\delta - \Lambda(T)) = 0.
$$ {#eq-MG-mean}
Thus if $\hat{\Lambda}_i$ were the true cumulative hazards function, 
the  martingale residuals would have mean zero. 

Now $\hat{\Lambda}_i$ is only an estimate, but if the model fits the data, the martingale residuals should have 
approximately mean $0$. In the more advanced theory of survival 
analysis they also enjoy a  martingale property, which explains their name. 
Here we will not pursue this, but just observe that we can investigate the model
fit by investigating if the martingale residuals  approximately have  mean $0$. 
The residuals can be used much as regression residuals to judge overall fit as
well as the fit of how the individual predictor variables enter into the model. 
It should be noted, though, that they by definition have a left skewed distribution[^mg-dis]
concentrated on $(-\infty, 1)$. 

[^mg-dis]: The martingale residuals are *never* approximately normally distributed!


::: {#exm-prostate-residuals}
As for other regression models, we can use `augment()` from the broom package 
to extract fitted values (the linear predictor) and residuals. @fig-prostate-mg-res-rx
shows the martingale residuals plotted against the treatment (which is the only 
predictor in the model at this point). Since the dataset is rather small, it makes 
sense here to use `geom_beeswarm()` from the ggbeeswarm package to visualize 
the full distribution of the residuals as well as the individual points.
Note the upper cap at $1$ of all martingal residual distributions.

```{r}
#| label: fig-prostate-mg-res-rx
#| fig-cap: "Martingale residuals for the Cox model for each of the four treatment groups. The blue points and errorbars are means and 95% confidence intervals."
#| fig-cap-location: margin
augment(prostate_cox, data = prostate) |> 
    ggplot(aes(rx, .resid)) + 
    ggbeeswarm::geom_beeswarm(alpha = 0.4) + 
    geom_point(
        stat = "summary", 
        fun = mean, 
        color = "blue", 
        size = 2.5
    ) +
    geom_errorbar(
        stat = "summary", 
        fun.data = mean_se, 
        fun.args = list(mult = 1.96),
        color = "blue", 
        width = 0.15, 
        linewidth = 0.7
    )
```

@fig-prostate-mg-res-rx includes means and confidence intervals for each 
group of residuals, which should include zero for all groups as in the figure. 

We can also plot the residuals against another potential predictor, such as `age`, 
as in @fig-prostate-mg-res-age. We see a clear association between age and the residuals 
suggesting that `age` (unsurprisingly) holds predictive information about 
the survival times. 

```{r}
#| label: fig-prostate-mg-res-age
#| fig-cap: "Martingale residuals for the Cox model against `age`, which was not included as a predictor in the model."
#| column: margin
#| opts.label: smallfigure
#| warning: false
#| echo: false
augment(prostate_cox, data = prostate) |> 
    ggplot(aes(age, .resid)) + 
    geom_point(alpha = 0.4) + 
    geom_smooth()
```

We want to include `age` in the model, and we will first try to include it as 
an additive term.

```{r}
#| label: prostate-cox-add
#| code-fold: show
prostate_cox_add <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ age + rx, 
    data = prostate, 
)
```

@fig-prostate-cox-add-res shows two residual plots. One based on comparing the 
Cox-Snell residuals to the exponential distribution and standard type of a residual 
plot, this time of the martingale residuals against the fitted values. The 
fitted values are the linear predictors. For the computation of the Cox-Snell 
residuals, we used that `type.predict = "expected"` in the call of `augment()`
gives $\hat{\Lambda}(T_i)$ as the predicted values (`.fitted` column in the output).

```{r}
#| label: fig-prostate-cox-add-res
#| fig-cap: "The Cox-Snell residual plot (left) and the martingale residuals plotted against fitted values (right) for the additive model."
#| fig-cap-location: margin
#| opts.label: widefigure2
#| message: false
p1 <- augment(prostate_cox_add, data = prostate, type.predict = "expected") |>
    survfit(Surv(.fitted, status != "alive") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) + 
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") + 
    ylab("Cumulative hazard")


p2 <- augment(prostate_cox_add) |> 
    ggplot(aes(.fitted, .resid)) +
    geom_point(alpha = 0.4) + 
    geom_smooth() + 
    xlab("Fitted values") +
    ylab("Martingale residuals")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

We can see minor model misspecifications from the residual plots in 
@fig-prostate-cox-add-res. First, there are deviations from a straight 
line in the Cox-Snell plot, suggesting some form of model misspecification. 
The smoothed curve in the martingale residual plot also suggests some 
minor model misspecification. We could suspect that `age` is not
correctly modeled.
:::

### Deviance residuals

Alternative residuals include the deviance residuals and the 
Schoenfeld residuals. We will not treat the latter, but the former are based on 
the following observation. The Poisson nonparametric log-likelihood as well as the parametric
log-likelihood with a continuous baseline can be written as 
$$
    \ell^* = \sum_{i=1}^n \underbrace{\delta_i \log (w_i \Lambda_{0,i}) - w_i
    \Lambda_{0,i}}_{\text{Poisson log-likelihood term}} + \sum_{i=1}^n \delta_i \log \left(\frac{\lambda_{0,i}}{\Lambda_{0,i}}\right)
$$ 
The first term is identical to the Poisson log-likelihood, and the
second term does not depend on parameters that enter into the weights. 
Based on this we define the deviance residuals in terms of the
Poisson deviances \index{residual!survival deviance}
$$
    d(\delta_i, \hat{\Lambda}_{i}) = 2\left( \delta_i \log
    (\delta_i/\hat{\Lambda}_{i}) - \delta_i + \hat{\Lambda}_{i}\right),
$$
where $\hat{\Lambda}_{i} =  \hat{w}_i \hat{\Lambda}_{0,i}$.

The result is the *deviance residuals*
$$
    \text{sign}(\delta_i - \hat{\Lambda}_i) \sqrt{d(\delta_i, \hat{\Lambda}_i)} = \left\{\begin{array}{ll}
    - \sqrt{2  \hat{\Lambda}_i} & \quad \text{if } \delta_i = 0 \\
    \text{sign}( 1 -  \hat{\Lambda}_i)\sqrt{ 2( \hat{\Lambda}_i -
    1  - \log \hat{\Lambda}_i)} & \quad
    \text{if } \delta_i = 1
    \end{array} \right.
$$ 

Observe that the deviance residuals can be expressed as
$$
    \text{sign}(\hat{r}_i) \sqrt{-2(\hat{r}_i + \delta_i \log (\delta_i - \hat{r}_i))}
$$
with $\hat{r}_i = \delta_i - \hat{\Lambda}_i$ the martingale
residuals. Compared to the martingale residuals, the distribution of 
the deviance residuals is less skewed.

### Prostate cancer continued

We conclude this section by a final analysis of the prostate cancer data
based on the insights we have obtained from the previous sections. 
We include both `age` and treatment `rx` in the model. Now with `age` 
nonlinearly expanded using natural cubic splines and `rx` interacting (linearly)
with `age`. 

```{r}
#| label: prostate-cox-int
#| code-fold: show
prostate_cox_int <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ ns(age, df = 4) + age * rx, 
    data = prostate, 
)
```

We first reconsider the residual plots in @fig-prostate-cox-int-res. In contrast to 
@fig-prostate-cox-add-res there are no indications in these residual plots of 
any model misspecification.

```{r}
#| label: fig-prostate-cox-int-res
#| fig-cap: "The Cox-Snell residual plot (left) and the martingale residuals plotted against fitted values (right) for the interaction model."
#| fig-cap-location: margin
#| opts.label: widefigure2
#| message: false
p1 <- augment(prostate_cox_int, data = prostate, type.predict = "expected") |>
    survfit(Surv(.fitted, status != "alive") ~ 1, data = _) |>
    summary(data.frame = TRUE) |>
    ggplot(aes(time, cumhaz)) + 
    geom_abline(slope = 1, color = "blue") +
    geom_point() +
    xlab("Cox-Snell residuals") + 
    ylab("Cumulative hazard")


p2 <- augment(prostate_cox_int) |> 
    ggplot(aes(.fitted, .resid)) +
    geom_point() + 
    geom_smooth() + 
    xlab("Fitted values") +
    ylab("Martingale residuals")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

To better understand the interaction model we will compute and visualize 
the estimated risk factors 
$$
    e^{\hat{\eta}}
$$
for some combinations of treatment and age as well as the corresponding fitted survival curves.

```{r}
#| label: prostate-pred-frame
#| code-fold: show
pred_frame <- expand.grid(
        age = c(55, 62.5, 70, 77.5),
        rx = factor(1:4, labels = levels(prostate$rx))
    )

pred_frame <- mutate(
        pred_frame,
        risk = predict(
            prostate_cox_int,
            newdata = pred_frame,
            type = "risk"
        ),
        risk = risk / risk[1],
        data = 1:16
    ) 
```



```{r}
#| label: tbl-prostate-pred-frame
#| tbl-cap: "Estimated risk factors for the interaction model."
#| tbl-cap-location: bottom
#| column: margin
#| echo: false
pred_frame |> knitr::kable(digits = 3)
```


The computations above give a data frame with estimated risk factors for 16 combinations 
of age and treatment. Since these factors are always relative risk factors, we have 
normalized them so that the placebo treatment for age 55 is the reference. @tbl-prostate-pred-frame
shows the table of the resulting risk factors. Recall that the smaller the risk 
factor is, the smaller is the risk of dying and the longer is the survival time. 


@fig-prostate-pred-frame shows a plot of the estimated risk factors, 
from which the nonlinear and non-additive nature of the model is 
clear. For the high treatment dosage there is a notable change from reducing the risk relative 
to placebo for patients younger than about 70 years while it increases the risk for the older patients.

```{r}
#| label: fig-prostate-pred-frame
#| fig-cap: "Estimated risk factors as a function of age and treatment relative to placebo treatment for patient of age 55."
#| fig-cap-location: margin
#| echo: false
pred_frame2 <- expand.grid(
        age = seq(55, 85, 1),
        rx = factor(1:4, labels = levels(prostate$rx))
    )

mutate(
    pred_frame2,
        risk = predict(
            prostate_cox_int,
            newdata = pred_frame2,
            type = "risk"
        ),
        risk = risk / risk[1],
    ) |>
    ggplot(aes(age, risk, color = rx)) + 
    geom_line(linewidth = 1) +
    theme(legend.position = "top")
```

@fig-prostate-int-surv shows the fitted survival functions corresponding 
to the risk factors in @tbl-prostate-pred-frame. It gives a different perspective 
on how the estimated survival curves change with age, and how the interaction 
effect makes the treatment with the high dosage appear beneficial for younger 
patients while harmful for older patients. 

```{r}
#| label: fig-prostate-int-surv
#| fig-cap: "Estimated survival functions for the Cox interaction model. The dashed curve is the Kaplan-Meier estimator for the entire data."
#| warning: false
#| message: false
#| column: page
#| opts.label: widefigure4
survfit(prostate_cox_int, newdata = pred_frame) |> 
    summary(data.frame = TRUE) |>
    left_join(pred_frame) |>
    ggplot(aes(time, surv, color = rx)) + 
    geom_step(
        data = summary(prostate_KM, data.frame = TRUE), 
        aes(color = NULL), 
        linetype = 2
        ) + 
    geom_step() +
    facet_wrap("age", ncol = 4) +
    theme(legend.position = "top")
```

We have not quantified the uncertainty in the above figures. One way to 
globally test if there is a treatment effect in this interaction model is 
via a likelihood-ratio test (using the partial likelihood). @tbl-prostate-int-anova
shows that the $p$-value is about $0.003$ and thus fairly small. Based on 
this test we therefore reject the hypothesis that there is no treatment effect. 

```{r}
#| label: tbl-prostate-int-anova
#| tbl-cap: "Likelihood-ratio test of the hypothesis that there is no treatment effect."
#| tbl-cap-location: bottom
prostate_cox_age <- coxph( 
    Surv(dtime + 0.5, status != "alive") ~ ns(age, df = 4),
    data = prostate, 
)
anova(prostate_cox_age, prostate_cox_int) |> 
    tidy() |> 
    knitr::kable()
```

Today, estrogen treatment of prostate cancer has largely been replaced 
by better alternatives. Though it is documented to be effective against the cancer, a 
serious known sideeffect is increased risk of cardiovascular disease. This 
knowledge is in line with our findings for this particular study. 

```{r}
#| eval: false
#| echo: false
coxph( 
    Surv(dtime + 0.5, status != "alive") ~ ns(age, df = 4) + age * hx * rx,
    data = prostate, 
) |> anova()
```

## Discrete time survival models {#sec-dis-surv}

We have in previous sections mostly considered survival distributions that 
are continuous, that is, distributions on $[0, \infty)$ with a continuous 
survival function $S$. It was typically even differentiable with 
$-S'$ the density of the distribution, in which case the hazard function 
is well defined. In this chapter we will briefly touch on discrete 
survival distributions, and almost exclusively on distributions with 
point masses in the observations $T_1, \ldots, T_n$. We assume throughout
this section that there are no ties.

::: {#def-emp-like-surv}
With observations $T_1, \ldots, T_n$ and $P$ a probability measure on $[0, \infty)$ 
with survival function $S$, the empirical likelihood with right censoring is 
$$
    L(P) = \prod_i P(T_i)^{\delta_i} S(T_i)^{1-\delta_i}.
$$
:::

The factors in the empirical likelihood are simply the probabilities
$$
    P(T_i)^{\delta_i} S(T_i)^{1-\delta_i} = \mathbf{P}(T^* = T_i, \delta = \delta_i) =
    \left\{ \begin{array}{ll} \mathbf{P}(T^* = T_i) & \quad \text{if } \delta_i = 1 \\
     \mathbf{P}(T^* > T_i) & \quad \text{if } \delta_i = 0 \end{array} \right.
$$

Recall also that for the discrete probability measure $P$, the survival function 
is 
$$
    S(t) = \mathbf{P}(T^* > t) = \sum_{i: T_i > t} P(T_i).
$$
We will show in this section that the Kaplan-Meier estimator
can be derived as the maximizer of the empirical likelihood.

### Discrete hazards

In the previous section we only introduced hazards for continuous distributions. 
The *discrete hazards* are by definition the conditional probabilities
$$
    \lambda_i \coloneqq \frac{P(T_i)}{S(T_i-)} = \mathbf{P}(T^* = T_i \mid T^* \geq T_i).
$$
Note that the discrete hazards are probabilities and thus constrained to lie in the interval
$[0,1]$.

::: {#lem-dissurv}
For a discrete probability measure with point masses in $T_1, \ldots, T_n$ and corresponding hazards
  $\lambda_i$ the survival function is
$$
    S(t) = \prod_{i: T_i \leq t} (1 - \lambda_i).
$$
:::

::: {.proof}
Note that $S(T_{(j)}-) = S(T_{(j-1)})$ for $j \geq 2$ and that it is sufficient to show the
identity for $t = T_{(j)}$, $j = 1, \ldots, n$. 

The proof is by induction. For $T_{(1)}$ we have $S(T_{(1)}-) = 1$ and 
$$
    S(T_{(1)}) = 1 - P(T_{(1)}) = 1 - \lambda_{(1)},
$$
which gives the induction start. 

For the induction step, observe that the definition of the hazard implies that
$$
    \begin{align*}
        S(T_{(j)})  & = S(T_{(j)}-) - P(T_{(j)}) \\
        & = S(T_{(j)}-) - S(T_{(j)}-)\lambda_{(j)} \\
        & = S(T_{(j-1)})(1-\lambda_{(j)}),
    \end{align*}
$$
which gives the induction step.
:::

@lem-dissurv shows that the discrete hazards determine the
survival function and thus the distribution, but in a slightly
different way than for continuous distributions. As a direct 
corollary we can express the empirical likelihood in terms of 
the discrete hazards.

::: {#cor-emp-like-surv-II} 
With $Y_i = Y(T_i) = \sum_{j} 1(T_i \leq T_j)$ the number of individuals at risk at time $T_i$,
the empirical likelihood equals
$$
    L(P) = \prod_{i} \lambda_i^{\delta_i} (1 - \lambda_i)^{Y_i - \delta_i}.
$$ {#eq-emp-like-surv}
:::

::: {.proof}
From the definition of the hazards and @lem-dissurv 
$$
    P(T_i) = \lambda_i S(T_i-) = \prod_{j: T_j \leq T_i} (1 - \lambda_j).
$$

We then plug this expression into the empirical likelihood function and interchange 
the product order to get
$$
    \begin{align*}
    L(P) & =  \prod_i \lambda_i ^{\delta_i}  \prod_{j: T_j < T_i} (1 -
        \lambda_j)^{\delta_i}  \prod_{j: T_j \leq T_i} (1 - \lambda_j)^{1-\delta_i} \\
    & =  \prod_{i} \lambda_i^{\delta_i} (1 - \lambda_i)^{Y_i - \delta_i}.
    \end{align*}
$$
:::

### Empirical likelihood estimators

In terms of the empirical likelihood we define the nonparametric maximum-likelihood 
estimator (NPMLE) of the survival distribution as the maximizer of $L$. To compute it we maximize 
$L$ as a function of the discrete hazards. To this end we need the following lemma.

::: {#lem-surv-opt}
With $y \geq 1$  the function 
$$
    \lambda \mapsto \lambda (1 - \lambda)^{y - 1}
$$
attains a unique maximum over $[0, 1]$ in $1/y$. 
:::

::: {.proof}
With $f(\lambda) =  \lambda (1 - \lambda)^{y - 1}$ we
  have
$$
    \begin{align*}
    f'(\lambda) & = (1 - \lambda)^{y - 1} - \lambda (y-1) (1 - \lambda)^{y - 2} \\
        & = (1 - \lambda)^{y - 2}(1 - \lambda - \lambda (y - 1)) \\
        & =  (1 - \lambda)^{y - 2} (1 - \lambda y).
    \end{align*}
$$
For $y = 1$, the function is the identity, which attains 
its maximum in $1 = 1/y$. For $y > 1$ we have $f(\lambda) \geq 0$ for $\lambda \in [0,1]$, 
$f(0) = f(1) = 0$ and the function has a 
unique stationary point in $(0, 1)$ in $1/y$ with $f(1/y) > 0$. 
Thus $1/y$ is the unique maximizer in $[0,1]$ for $y \geq 1$. 
:::

::: {#thm-npmle-surv}
The NPMLE of the hazards -- the hazards for the
  probability measure that maximizes the empirical likelihood -- is 
given by
$$
    \hat{\lambda}_i = \frac{\delta_i}{Y_i}.
$$
::: 

::: {.proof}
The factors of $L$ in ([-@eq-emp-like-surv]) equal
$$ 
    \lambda_i^{\delta_i} (1 - \lambda_i)^{Y_i - \delta_i}
    = \left\{ \begin{array}{cc}
    (1-\lambda_i)^{Y_i} & \quad \text{if } \delta_i = 0 \\
    \lambda_i (1-\lambda_i)^{Y_i - 1} & \quad \text{if } \delta_i = 1 \\
    \end{array} \right.
$$
In the first case, when $\delta_i = 0$, the factor is maximized in
$0$, and in the second case, when $\delta_i = 1$, the factor 
is maximized in $1/Y_i$. 
::: 


By plugging in the NPMLE of the hazards in the formula in @lem-dissurv for the survival function we
find that the NPMLE of the survival function is 
$$
    \hat{S}(t) = \prod_{i: T_i \leq t} \left(1 - \frac{\delta_i}{Y_i}\right).
$$
We recognize this as the Kaplan-Meier estimator. 

In terms of the discrete hazards the corresponding jump cumulative 
hazard function is
$$
    \Lambda(t) = \sum_{i : T_i \leq t} \lambda_i.
$$
If we plug in the NPMLE of the hazards in this formula, we get 
$$
    \hat{\Lambda}(t) = \sum_{i : T_i \leq t} \frac{\delta_i}{Y_i},
$$
which we recognize as the Nelson-Aalen estimator. The previously observed  
discrepancy between the Kaplan-Meier and Nelson-Aalen estimators 
was due to the relation in ([-@eq-surv-by-NA]) assuming a continuous distribution. 

The two main points from this section are:

* The Kaplan-Meier estimator has a justification as a nonparametric maximum-likelihood estimator
* When we use the *correct* relation between hazards and the survival function for discrete 
    distributions, as given by @lem-dissurv, there is no discrepancy between the 
    Kaplan-Meier and Nelson-Aalen estimators. 

### The discrete proportional hazards model

It is legitimate to consider the proportional hazards model
for discrete probability distributions, and use the 
empirical likelihood function $L$ for profiling. Contrary 
to the derivations in @sec-pois-emp-like, that use the Poisson empirical likelihood, 
the maximizer is not explicit. Moreover, the proportionality assumption for the 
discrete model does not really provide an approximation of the 
proportionality assumption for the continuous model. 

The appropriate generalization of the proportional hazards model 
to the discrete case is actually the model
$$
\lambda_i = 1 - (1 - \lambda_{0,i})^{w_i}.
$$ {#eq-disprop}

The empirical likelihood becomes \index{empirical likelihood!discrete survival}
$$
    L = \prod_{i} ( 1 - (1 - \lambda_{0,i})^{w_i})^{\delta_i} (1 -
    \lambda_{0,i})^{W_i - w_i \delta_i},
$$
and similar arguments as in the proof of @thm-npmle-surv will
show that the NPMLE of $\lambda_{0,i}$ for given weights is 
$$
    \hat{\lambda}_{0,i} = \delta_i\left(1 - \left(1 - \frac{w_i}{W_i}\right)^{1/w_i}\right).
$$
The corresponding profile likelihood is 
$$
    \prod_{i: \delta_i = 1} \frac{w_i}{W_i}  \left(1 - \frac{w_i}{W_i}\right)^{W_i/w_i - \delta_i}.
$$ {#eq-npprof}

The discrete proportional hazards model ([-@eq-disprop]) is arguably
the "correct" generalization of the continuous proportional
hazards model, see @Jacobsen:1984, and using the empirical likelihood yields
the profile likelihood ([-@eq-npprof]). The resulting profile likelihood is not Cox's partial
likelihood, but it is for most practical purposes very close. In the
literature, Cox's partial likelihood has caught on as the default
choice for semiparametric estimation of parameters that enter the
weights. 

## Exercises {.unnumbered}

::: {#exr-loglog}
Show that if $T$ has a log-logistic distribution with density $f = f_0$ then 
$$
    e^{\eta} T
$$ 
has density $f_{\eta}$. Show that if $T$ has a log-logistic distribution with density
$f_{\eta}$ then 
$$
    \log T - \eta
$$
has a logistic distribution. 
::: 

::: {#exr-empsurv}
Show that without censoring then the Kaplan-Meier estimator 
equals the empirical survival function \index{empirical survival function}
$$
    \hat{S}(t) = \frac{1}{n} \sum_{i=1}^n 1(T^*_i > t).
$$
::: 

::: {#exr-prostate}
In this exercise you should consider the prostate dataset from the RwR package
and this time include the variable `hx` in the analysis. The variable `hx`
indicates if the patient has a history of cardiovascular disease. 
You should ignore the treatment variable `rx`.

1. Fit a proportional hazards model using the two variables `age`
    and `hx` as predictors. 
2. Make one or more residual plots for the model you have fitted. Does it fit
    the data?
3. Test the hypothesis that survival is not associated with `hx` given `age`.
:::

::: {#exr-weibull}
Let $T_1^*, \ldots, T^*_n$ be independent survival times for $n$ patients and let
$X_i \in \{0, 1\}$ indicate if the $i$-th patient is treated
$(X_i = 1)$ or untreated $(X_i = 0).$ We are interested in estimating the
treatment effect, which we summarize in the following *parameter of interest*
$$
    \rho := \frac{\text{median}(T_1^* \mid X_1 = 1)}{\text{median}(T_1^* \mid X_1 = 0)},
$$
that is, $\rho$ is the ratio of the median survival time for 
treated vs. untreated patients.

In this exercise it is assumed that $T_i^*$ follows a Weibull
distribution with shape parameter $\gamma > 0$ and scale parameter
$\alpha_{x_i}$, that is, the hazard rate is 
$$
    \lambda_i(t) = \alpha_{x_i} \gamma t^{\gamma-1}.
$$

1. Show that 
    $$
        \rho = \sqrt[\gamma]{\frac{\alpha_0}{\alpha_1}}
    $$
    in terms of the Weibull parameters.

2. Show that the Weibull model can be formulated as a proportional
    hazard model in terms of the linear predictor 
    $$
        \eta_i = \beta_0 + \beta_1 X_i
    $$
    and relate the parameters $\beta_0$ and $\beta_1$ to $\alpha_0$ and
    $\alpha_1$. Express $\rho$ in terms of the $\beta_0$ and $\beta_1$ parameters.

In the remaining part of this exercise we will study
two different estimators of $\rho$ on simulated data. To this end introduce
i.i.d. censoring times $C_1, \ldots, C_n$ and the censored observed
survival times 
$$
    T_i = \min\{T_i^*, C_i\}.
$$

The data set `testWei` available from the R package RwR is a simulated data
set with $n = 200$, $X_1 = \ldots = X_{100} = 0$ and $X_{101} = \ldots = X_{200} = 1$, 
with $\gamma = 2$, $\beta_0 = 0$ and $\beta_1 = -1$ (so that $\rho \approx 1.649$). 
The censoring times are uniformly distributed on $[0, 4]$. The `time` column
contains the $T_i$-s and the `status` column is the indicator
  of censoring ($1$ is uncensored, $0$ is censored). 

3. Compute and plot the Kaplan-Meier estimates of the survival
    function within each of the two treatment groups for the simulated `testWei` data set.

4. Use the Weibull model with fixed $\gamma = 2$ to estimate $\beta_0$
    and $\beta_1$. Compare the resulting estimated survival functions from
    the Weibull model with the Kaplan-Meier estimates.

5. Compute the MLE $\hat{\rho}_{\mathrm{MLE}}$ of $\rho$ based on the Weibull model
    (with fixed $\gamma = 2$) and compute a 95\% confidence interval for
    $\rho$. Compute also, based on the Kaplan-Meier estimates, a direct
    estimate, $\hat{\rho}_{\mathrm{KM}}$, of $\rho$. 

6. Write an R program that simulates a data set (including the censoring) using the same 
    setup as for the `testWei` data set described above. Simulate $B = 200$
    data sets and compute for each data set the estimates $\hat{\rho}_{\mathrm{KM}}$ and
    $\hat{\rho}_{\mathrm{MLE}}$ as above and compare their distributions. 

*Hint: Use the R function `rweibull()` to simulate from the Weibull distribution, but pay attention to the parametrization.*
:::

::: {#exr-lung}
For this exercise we consider the data set `lung` on lung cancer survival 
available from the R package RwR. It contains four variables:

| Variable | Description                      |
|----------|----------------------------------|
| time     | survival time in days            |
| status   | censored = 1, dead = 2           |
| age      | age in years                     |
| sex      | male = 1, female = 2             |

1. Plot the Kaplan-Meier estimates of the survival functions for males and females separately.
2. Use the Cox proportional hazard model to investigate if the survival time depends upon age. 
3. Check if the Cox proportional hazard model fits the data. 
::: 

::: {#exr-prop-haz}
This exercise is about the proportional hazard model with cumulative hazard function 
$$
    \Lambda_i(t) = e^{\eta_i} \Lambda_0(t)
$$
for the $i$-th individual and linear predictor 
$$
    \eta_i = \beta_0 + \beta_{\texttt{age}} X_{i, \texttt{age}} + \beta_{\texttt{sex}} X_{i, \texttt{sex}}.
$$
Thus the two predictor variables are the age and sex of the individual. 

You will need that data set `simSurv` from the R package RwR that contains 
the four variables, `surv`, `status`, `age` and `sex` being the censored survival times, the
censoring indicator, the age and the indicator of gender (1 is male), respectively, of the
individuals. There is a total of 200 individuals.

1. Estimate, using Cox's proportional hazard model, the parameters $\beta_{\texttt{age}}$
    and $\beta_{\texttt{sex}}$. Compute a 95\% confidence interval for $\beta_{\texttt{sex}}$ and determine 
    whether males have a higher risk of dying than females. 

For the remaining questions the baseline cumulative hazard function is assumed to be
$$
    \Lambda_0(t) =  e^{\gamma t} - 1, \quad t \geq 0.
$$
The cumulative hazard function for the $i$-th individual is thus
$$
    \Lambda_i(t) = e^{\eta_i} ( e^{\gamma t} - 1).
$$

We assume that we have independent censored observations $(T_1, \delta_1), \ldots, (T_n, \delta_n)$ with 
$T_i \geq 0$ the $i$-th censored survival time and $\delta_i$ the indicator of whether the $i$-th observation is censored.

2. Compute the baseline hazard rate $\lambda_0(t)$ and show that the log-likelihood function is 
$$
    \sum_{i} \delta_i \eta_i + \delta_i \log (e^{\gamma T_i} - 1) - e^{\eta_i} (e^{\gamma T_i} - 1) + 
    \delta_i \log \frac{\gamma e^{\gamma T_i}}{e^{\gamma T_i} - 1}.
$$

3. Plot the profile log-likelihood for the $\gamma$-parameter. Compute the MLE of $\gamma$, $\beta_0$,
  $\beta_{\texttt{age}}$ and $\beta_{\texttt{sex}}$. 

*Hint: Identify the first two terms of the log-likelihood as the log-likelihood for a Poisson generalized linear model. Use the `glm()`-function for the profiling.*

4. Use the glm framework to compute a 95\% confidence interval for $\beta_{\texttt{sex}}$ by plugging in the MLE,
    $\hat{\gamma}$, of $\gamma$ and pretending that it is fixed and known. Compute next a 95\% confidence interval using non-parametric bootstrapping. Compare the results. 
::: 

::: {#exr-cox-alg}
This exercise deals with the numerical algorithms associated with the optimization of Cox's partial likelihood. 
Consider censored survival data $(T_i, \delta_i, X_i)$ for $i = 1, \ldots, n$ 
with $T_i$ the possibly censored survival time, $\delta_i$ the indicator of whether
$T_i$ is uncensored, and $X_i \in \mathbb{R}$ a univariate real valued
predictor. The proportional hazard model is given by the individual weights 
$$
    w_i = e^{\beta X_i}
$$
for an unknown parameter $\beta$. In the following we will make the assumption that the survival 
times are ordered and with no ties, that is
$$
    T_1 < T_2 < \ldots < T_n.
$$

1. Show that Cox's partial log-likelihood equals 
    $$
        \ell(\beta) = \sum_{i: \delta_i = 1} \beta X_i - \log \left(\sum_{j \geq i} e^{\beta X_j} \right),
    $$
    and compute the first and second derivative of $\ell$ w.r.t. $\beta$.

The estimator $\hat{\beta}$ is the maximizer of $\ell$ provided that $\ell$ attains a maximum. It then solves the equation
$$
    \ell'(\beta) = 0.
$$

2. Show that $\ell$ is concave. Find a condition on the $X_i$-s that ensure that $\ell$ has a unique maximum. 

For the remaining part of this problem you first need to 
simulate a data set from the model above. You may experiment 
with different choices of $\beta$, but try also $\beta = 0$. 

3. Implement Newton's algorithm for computing $\hat{\beta}$.
    Apply the algorithm with start value $\beta_{1} = 0$ to the simulated
    data set and investigate the values of $\beta_m$, $\ell(\beta_m)$,
    $\ell'(\beta_m)$ and $\ell''(\beta_m)$ in the iterations of the algorithm.

[A naive implementation of the computation of e.g. $\ell$ scales like $n^2$. 
 Make sure that your implementation scales like $n$.]{.aside}

4. Compute and plot an estimate of the baseline cumulative hazard function.

5. Compute $\hat{\beta}$ using the `coxph()` function in R. Compute also a standard 95\% confidence interval for
    $\beta$. Compare this interval with a (partial) likelihood interval based on $\ell$. 
:::

::: {#exr-mg-res}
This exercise is on theoretical properties of the survival analysis martingale
residuals. Let $T^*$ and $C$ be independent random variables on $[0,\infty)$, whose
distributions have continuous densities $f$ and $g$, respectively. The
corresponding survival functions are denoted $S$ and $H$. Note that $f
= - S'$ and $g = - H'$. Then $T = \min\{T^*, C\}$ denotes the observed
(censored) survival time and $\delta = 1(T^* \leq C)$ is the status
indicator. 

1. Show that for any function $K: [0,\infty) \times \{0,1\} \to \mathbb{R}$, continuously
    differentiable in the first variable, it holds that 
    $$
        \begin{align*}
        \mathbf{E}( K(T, \delta)) & = \int_0^\infty \Big(K(0,1) + (K(s, 0) -  K(s,1)) S(s) \\
        & \quad + \int_0^s \partial_tK(t, 1) S(t) \mathrm{d} t \Big) g(s) \mathrm{d} s
        \end{align*}
    $$
    provided that the expectation is well defined. Use this to show that[^known] 
    $$
        \mathbf{E} (\delta - \Lambda(T)) = 0 \qquad \text{and} \qquad \mathbf{V}(\delta - \Lambda(T)) = \mathbf{E}(\Lambda(T))
    $$
    where $\Lambda$ denotes the cumulative hazard function for the distribution given by $f$. 

[^known]: That the expectation is $0$ is derived in the main text, but a slightly different derivation is provided here. The variance formula is new.

Recall that for the proportional hazard model, the $i$-th
individual has cumulative hazard function $w_i \Lambda_0$ where $w_i$
is an individual weight and $\Lambda_0$ is the baseline hazard. 

Assume
that $\hat{w}_i$ is the estimate of the $i$-th weight and that
$\hat{\Lambda}_0$ is the estimate of the baseline. With
$\hat{\Lambda}_i = \hat{w}_i \hat{\Lambda}_0(T_i)$ the fitted
cumulative hazard function for the $i$-th individual evaluated in $T_i$, 
$$
    \hat{r}_i = \delta_i - \hat{\Lambda}_i
$$
is the $i$-th martingale residual. Let $\hat{d}_i$ denote the $i$-th deviance residual[^dev-res]. 

2. Give an argument for considering the following standardization of the martingale residuals:
    $$
        \tilde{r}_i = \frac{\hat{r}_i}{\sqrt{\hat{\Lambda}_i}}.
    $$
3. Show that if $\delta_i = 1$ then 
    $$
        \hat{d}_i^2 = \tilde{r}_i^2 + o((1 - \hat{\Lambda}_i)^2).
    $$ 
4. Explore graphically how good the approximation
    $$
        \hat{d}_i \approx \tilde{r}_i
    $$
    is as a function of $\hat{\Lambda}_i$ when $\delta_i = 1$. Does a similar approximation hold if $\delta_i = 0$?

[^dev-res]: The deviance residuals are sometimes regarded as distribution stabilized transforms of the martingale residuals. This
exercise gives some insight on this perspective. 
::: 
