::: {.hidden}
{{< include _macros.qmd >}}
:::

```{r}
#| echo: false
#| cache: false
source("_common.R")
```


# Regression modeling {#sec-regression-modeling}

::: {.callout-important}
This chapter is currently undergoing conversion and substantial changes 
will be made without notice. For the sections on practical data analysis 
we refer, for the time being, to the 2020-edition.
:::

This chapter deals with a range of general problems and questions related to 
regression modeling and the practical process of data analysis. Some of the general 
considerations apply to statistics and data analysis in a broad sense, but to keep focus, 
data analysis is treated in the framework of regression models only. 

In the first part of the chapter we position the linear model from 
@sec-linear within a more general framework. This makes it possible to 
formulate some of the central objectives in regression analysis without 
relying on specific model assumptions. 

In the subsequent parts we turn to the process of practical data analysis, 
which involves a lot of decision making. For an 
inexperienced data analyst the decision making can be a daunting task, first of all 
because there is never a set of clearly correct decisions.  Instead it appears 
as if many decisions have to be made on the basis of little knowledge, and whenever 
a decision is made, it may be open to criticism. It is therefore tempting to 
hide the decision process, but this can be detrimental to the scientific 
process.  

We will develop methods for understanding and controlling the decision making 
process involved in practical data analysis. This ranges from grand decisions, 
like what the overall purpose of the analysis is, to technical decisions, 
such as how to handle missing values. Central to the decision process is a 
a solid understanding of how we can interpret an associational regression 
model and what it can ultimately be used for. For this reason we will also 
briefly treat some central *causal* questions and how and when we can use 
regression models to answer them.

The chapter concludes in @sec-nonlinear-expansions by introducing basis expansion 
techniques to capture nonlinear relations within the framework of the linear model.
The techniques are illustrated by a spline basis expansion of the relation 
between claim size and insurance sum in the fire insurance example.   

## Outline and prerequisites


The chapter primarily relies on the following four R packages. 

```{r}
#| label: setup
#| cache: false
#| code-fold: false
library(RwR)
library(ggplot2)
library(tibble)
library(broom)
library(splines)
library(dplyr)
```


## General regression models

There exists a range of generalizations of the linear regression model. It would be 
misleading to lump them all together under the umbrella term *non*linear regression. 
Some generalizations focus specifically on nonlinear relations between the response and 
the predictors, while other generalizations focus on other targets than the conditional 
expectation, e.g., the median or certain quantiles. 

We will consider various generalizations in this book. In 
@sec-generalized-linear the target is still 
the conditional expectation -- with a particular form of nonlinearity. But just as importantly, 
the chapter generalizes both the estimation methodology and the statistical theory to account 
for variance heterogeneity. 

In @sec-survival the target is the conditional hazard function of a 
survival distribution. One particular class of such survival regression models are
models of the conditional expectation of the *log-survival time*, but most models are not 
directly targeting a conditional expectation.  

In this chapter we treat some general aspects of regression modeling
without making any of the linear modeling assumptions A1, A2 or A3, but 
for clarity, the modeling target will remain the conditional expectation throughout.

::: {#def-regression-function}
Let $X \in \mathbb{R}^p$ and $Y \in \mathbb{R}$ be random variables and suppose that $\mathbf{E}(|Y|) < \infty$. 
The regression function $m : \mathbb{R}^p \to \mathbb{R}$ is defined by 
$$
    m(x) = \mathbf{E}(Y \mid X = x)
$$ {#eq-regression-fct}
:::

The regression function $m$ is in general only uniquely defined almost surely w.r.t. to the 
distribution of $X$. Though if it has a continuous version, that version is unique on 
the support of $X$. In most applications it is reasonable to assume that $m$ is continuous.

A general regression model is a set $\mathcal{M}$ of regression functions. That is, 
$$
    \mathcal{M} \subseteq \{ m \mid m : \mathbb{R}^p \to \mathbb{R} \}.
$$
We see that the linearity assumption A1 can be expressed in terms of the regression function as 
$$
    m(x) = x^T \beta,
$$
which means that the corresponding linear regression model can be expressed as 
$$
    \mathcal{M}_{\mathrm{linear}} = \{ x \mapsto x^T \beta \mid \beta \in \mathbb{R}^p \}.
$$
The linear regression model is parametrized by the parameter $\beta$ belonging to the parameter set $\mathbb{R}^p$. 

Most regression models are parametrized by a parameter, abstractly denoted $\theta \in \Theta$, 
and we then write the model as 
$$
    \mathcal{M} = \{m_\theta \mid \theta \in \Theta \}.
$$

::: {#exm-logistic-curve}

## Dose-response curves

A simple classical example is the *logistic function* for $p = 1$: 
$$
    m_{(a,b,c)}(x) = \frac{\beta}{1 + e^{-(b x + a)}} + \alpha
$$ {#eq-logistic-curve}
The parameter $\theta = (a, b, \alpha, \beta)$ is $4$-dimensional and the parameter set is 
$\Theta = \mathbb{R}^4$. 
The S-shaped graph of a logistic function can, for instance, model the relation
between a dosage ($x$) and a continuous response ($y$). 
:::

::: {#exm-neural-network}

## Neural networks

Let 
$$
    \mu, h: \mathbb{R} \to \mathbb{R}
$$ 
be two fixed (nonlinear) functions. 
With $\mathbf{B}$ a real $p \times q$ matrix, $\beta \in \mathbb{R}^{q}$ and $\alpha \in \mathbb{R}$
we define the regression function
$$
    m_{\mathbf{B}, \alpha, \beta} (x) = \mu(\alpha + h(x^T \mathbf{B}) \beta)
$$
parametrized by $(\mathbf{B}, \alpha, \beta)$. The convention above is that $h$ is 
applied coordinatewisely to the $q$-dimensional (row)-vector $x^T \mathbf{B}$.

[The functions $h$ and $\mu$ are known as *activation functions* 
in the literature on neural networks.]{.aside}
This is an example of a neural network with a single hidden layer. The 
matrix $\mathbf{B}$ parametrizes the *input layer* mapping the predictor 
vector $x \in \mathbb{R}^p$ to $h(x^T \mathbf{B}) \in \mathbb{R}^{q}$. It is 
a composition of the parametrized linear map $x \mapsto x^T \mathbf{B}$ and a fixed coordinatewise 
nonlinear function $h$. The number $\alpha$ and the vector $\beta$ parametrize the *output layer*, which is 
likewise a composition of a linear map and the nonlinear function $\mu$. 

We introduce the notation $L_{\alpha, \mathbf{B}}^{h} (x) = h(\alpha + x^T \mathbf{B})$ for an activation
function $h$, a $p \times q$ matrix $\mathbf{B}$ and a vector $\alpha \in \mathbb{R}^q$. 
Then $L_{\alpha, \mathbf{B}}^h : \mathbb{R}^p \to \mathbb{R}^q$ 
and it represents a single layer in a neural network. Using this notation we 
can write the neural network regression function above as a composition of two such maps:
$$
    m_{\mathbf{B}, \alpha, \beta} = L_{\alpha, \beta}^\mu \circ L_{0, \mathbf{B}}^{h}
$$

A general neural network with $k$ hidden layers can be defined as
$$
    \begin{align*}
    m_{\alpha_1, \mathbf{B}_1, \ldots, \alpha_k, \mathbf{B}_k, \alpha, \beta}
    %& = \mu(h_k( \cdots ( h_2(h_1(x^T \mathbf{B}_1)\mathbf{B}_2 + \beta_2) \cdots \mathbf{B}_k) + \beta_k)\beta + \beta_0) \\
    & =  L_{\alpha, \beta}^\mu \circ L_{\alpha_k, \mathbf{B}_k}^{h_k} \circ  \ldots \circ L_{\alpha_1, \mathbf{B}_1}^{h_1} 
    \end{align*}
$$
The model is parametrized by $(\alpha_1, \mathbf{B}_1, \ldots, \alpha_k, \mathbf{B}_k, \alpha, \beta)$ where 
$\mathbf{B}_{j}$ is a $p_{j-1} \times p_j$ real matrix ($p_0 = p$),
$\alpha_j \in \mathbb{R}^{p_j}$, $\beta \in \mathbb{R}^{p_k}$ and $\alpha \in \mathbb{R}$. 

It is quite common to use the same activation function for all layers, 
that is, $h_1 = h_2 = \ldots = h_k = \mu$. A couple of standard choices are:
$$
    \begin{align*}
    \mu(\eta) & = (1 + e^{-\eta})^{-1} \qquad \mathrm{(logistic)} \\
    \mu(\eta) & = \log(1 + e^\eta) \qquad \ \mathrm{(softplus)} \\
    \mu(\eta) & = \max(0, \eta) \qquad \ \ \ \mathrm{(ReLU)}
    \end{align*}
$$
[The $4$-parameter dose-response model above is an example of a neural 
network with $p = q = 1$ and activation function being the logistic
function.]{.aside}

We will in @sec-generalized-linear treat *generalized linear models*, 
which can be regarded as neural networks with no hidden layers. That is, 
the regression function is of the form
$$
    m_\beta(x) = \mu(x^T \beta)
$$
We will in that context see canonical choices of the activation function $\mu$ 
dictated by the distribution of the response variable. 
:::

Even though the parameter space of a neural network can be very large -- in 
particular for deep neural networks with many hidden layers -- the parameter 
space is formally a finite dimensional space. We might 
want to choose a model $\mathcal{M}$ that is even larger, e.g., the set of all continuous functions, 
the set of all Lipschitz functions, or the set of all smooth functions. These sets 
cannot be parametrized in a nice way by a finite dimensional parameter[^parametric], 
and we call the corresponding regression models *nonparametric*. 

[^parametric]: More precisely, they are not finite dimensional differentiable 
manifolds.

There are substantial theoretical differences between trying to estimate a 
regression function in a parametric model or in a nonparametric model. 
From a practical perspective, though, the transition 
from small parametric models to nonparametric models via large parametric 
models is much more gradual. We will 
not deal with estimation methods or theory that are explicitly nonparametric, but 
it can be conceptually advantageous to define and discuss the 
purposes and objectives of a regression analysis in a nonparametric way. 
Doing so, we will be able to define our aims clearly without reference to a 
particular parametric model -- even though we eventually, in this book, 
develop solutions in terms of specific parametric models. 
 
### Estimation methodologies

The most widely used estimator of the regression function is the least squares 
estimator minimizing the squared error loss. 
With a model $\mathcal{M}$ and observations $(X_1, Y_1), \ldots, (X_n, Y_n)$ 
it is formally defined as 
$$
    \hat{m} = \argmin_{m \in \mathcal{M}} \sum_{i=1}^n (Y_i - m(X_i))^2.
$$ {#eq-nonlinear-least-squares}
If the model is parametrized by $\theta \in \Theta$, we can formulate 
the estimator in terms of estimating $\theta$
$$
    \hat{\theta} = \argmin_{\theta \in \Theta} \sum_{i=1}^n (Y_i - m_{\theta}(X_i))^2.
$$ {#eq-nonlinear-least-squares-theta}

::: {#exm-birth-weight-nn-fit}

## Birth weight

:::


[For sufficiently flexible models $\mathcal{M}$, e.g., 
models containing all degree $n$ polynomials, there is 
a subset $\mathcal{M}_0 \subseteq \mathcal{M}$ such that 
for all $m \in \mathcal{M}_0$, $m(X_i) = Y_i$ for all $i = 1, \ldots, n$. 
All regression functions in $\mathcal{M}_0$ are said to *interpolate* 
the data perfectly, and they all achieve the globally minimal squared error loss of $0$.]{.aside}
There are several practical problems with these formal definitions, though. 
First, there may be no global minimizer of the sum of squares or there may 
be many. Second, it may be practically impossible to compute any such global 
minimizer unless $\mathcal{M}$ is particularly nice, e.g.,
$\mathcal{M} = \mathcal{M}_{\mathrm{linear}}$ as in @thm-normal. Third, 
even if we could compute $\hat{m}$ it might not be a good estimator 
if $\mathcal{M}$ is very large relative to the sample size $n$. 

Many practical estimation algorithms do not attempt to litterally 
compute $\hat{m}$ or $\hat{\theta}$ as defined by ([-@eq-nonlinear-least-squares])
and ([-@eq-nonlinear-least-squares-theta]), respectively. Typically, 
they iteratively find and improve candidate regression functions in the model
-- using the squared error loss as guide rather than 
trying to drive it to its minimal value. This can be done in many ways, 
e.g., by explicitly restricting the model space to an $n$-dependent subset 
$\mathcal{M}_n \subseteq \mathcal{M}$, or by regularizing the estimator 
via a penalty function $\mathrm{pen} : \mathcal{M} \to \mathbb{R}$. 
In the latter case, the algorithm would then (try to) compute   
$$
    \hat{m}^{\lambda_n} = \argmin_{m \in \mathcal{M}} \sum_{i=1}^n (Y_i - m(X_i))^2 + \lambda_n \mathrm{pen}(m)
$$ {#eq-nonlinear-least-squares-pen}
for an $n$-dependent parameter $\lambda_n > 0$ controlling the tradeoff 
between the squared error loss and the penalty term. Typically, 
$\lambda_n \to 0$ for $n \to \infty$ so that the regularization 
has less and less impact on the estimator as the sample size increases.
@sec-nonlinear-expansions demonstrated by example the use of penalization 
within the framework of a linear model, where the penalization was used to 
control the fluctuations of nonlinear interaction terms. 

We can interpret the term $\lambda_n \mathrm{pen}(m)$ as an *inductive bias*, 
where we bias the estimator toward small values of $\mathrm{pen}(m)$. Another 
way to introduce an inductive bias is by *early stopping*. If we start the 
optimization algorithm in a particular subset $\mathcal{M}_0$ and then 
stop it before convergence, the result is implicitly biased toward models 
in $\mathcal{M}_0$. An algorithm could implement early stopping by monitoring
the squared error loss on an *independent dataset* and stop when that error 
does not improve anymore. Similarly, the parameter $\lambda_n$ could be selected 
by minimizing the squared error loss on an independent dataset over 
a range of possible values of $\lambda_n$. 
[We use *inductive bias* to mean a model assumption or algorithmic
constraint that stear the estimator in a particular direction. 
The concept is philosophically close to a Bayesian prior, and 
Bayesian statistics offers a principled way to transform prior 
beliefs to inductive biases. The folklore interpretation of 
the *No Free Lunch theorem* is that we cannot estimate $m$ 
to generalize beyond data without some inductive bias.]{.aside}

For a notable part of this book we consider parametrized models and 
loss functions that are nice enough for the estimators we consider 
to actually be global minimizers, such as $\hat{\theta}$ in 
([-@eq-nonlinear-least-squares-theta]). Moving beyond those models,
regression function estimators used in practice are mostly the result of running 
a complicated algorithm on data, which does not have a simple analytic 
representation as, e.g., a minimizer or a solution to an equation. 
This does not make such estimators bad, but it does make it somewhat 
more complicated to understand their theoretical properties. The 
least squares estimator in $\mathcal{M}_{\mathrm{linear}}$, as 
treated in detail in @sec-linear, stands out as having a 
particularly well understood theory, which we can leverage for 
practical statistical data analysis. We cannot obtain a 
similarly simple theory for general regression function estimators, but 
we formulate our modeling objectives in terms of general regression 
functions in clear and simple way that does not hinge 
on simplistic model assumptions.  

### Regression modeling objectives

As discussed at some length in the introduction, @sec-introduction, one purpose of 
estimating the regression function $m$ is to use $\hat{m}$ to make predictions 
about $Y$ from observation of $X$. This is the primary focus in most of the machine learning 
literature. Though we will cover this aspect in later chapters -- where we discuss 
quantification and optimization of predictive strength of a model -- we also deal at length 
with other objectives. 

[In this section we focus for clarity on how $m$ depends on a single coordinate. 
Similar questions can, of course, be formulated for any group of coordinates.]{.aside}
One main question we consider is:

**Does $m$ depend on $x_i$?**

To formulate this precisely, let $x_{-j} = (x_1, \ldots, x_{j-1}, x_{j+1}, \ldots, x_p)^T$ denote 
the $(p-1)$-dimensional vector with the $j$-th coordinate removed from $x$, and let
$$
    m_{-j}(x_{-j}) = \mathbf{E}(Y \mid X_{-j} = x_{-j})
$$
denote the regression function based on the $X_{-j}$-predictors only. We can then 
formulate the hypothesis 
$$
    H_0: \mathbf{E}(Y \mid X) = m_{-j}(X_{-j}).
$$ {#eq-fund-hyp}
This hypothesis means that $m$ does not depend on the $i$-th coordinate,
and it is an example of a conditional (mean) independence hypothesis.
We will develop statistical theory to test $H_0$ based on various regression 
modeling frameworks. 

Another main objective we consider is to:

**Quantify (and estimate) how much $m$ depends on a particular coordinate $x_i$.**

Obviously, this objective is closely related to the hypothesis 
above -- any sensible quantification should show that $m$ does not depend on $x_i$
if $H_0$ is true. Additionally, such a quantification is naturally tied together
with the development of a test statistic of $H_0$. Testing $H_0$ is, however, 
a simpler problem, and the challenge is to develop easily interpretable 
quantifications that are simultaneously easy to estimate. The regression modeling 
frameworks we consider will have some natural ways of quantifying such 
conditional associations.  

To aid the general discussion of the two objectives above, we introduce 
partial dependence functions and partial errors below. 
To ease notation we will use the convention that
$$
    m(x_j, x_{-j}) = m((x_1, \ldots, x_{j-1}, x_j, x_{j+1}, \ldots, x_{p})^T),
$$
where $m(x_j, x_{-j})$ is simply a shorthand notation for the right hand side 
above. Recall that $m(x) = \mathbf{E}(Y \mid X = x)$ denotes the joint regression function 
of $Y$ on $X$.

::: {#def-partial-dependence} 
The partial dependence function $m_j : \mathbb{R} \to \mathbb{R}$ for the $j$-th coordinate is 
$$
    m_j(x_j) = \mathbf{E}(m(x_j, X_{-j})) = \mathbf{E}(\mathbf{E}(Y \mid  X_j = x_j, X_{-j})).
$$ {#eq-partial-dependence}
::: 

The partial dependence function is also known as the adjusted regression function. 
It is essential for its definition that we integrate out over the *marginal* 
distribution of $X_{-j}$ and not the conditional distribution of $X_{-j}$ given 
$X_{j} = x_j$.  


Recall that $m_{-j}(x_{-j}) = \mathbf{E}(Y \mid X_{-j} = x_{-j})$ is the regression function 
of $Y$ on $X_{-j}$ excluding the $j$-th coordinate of $X$.

::: {#def-partial-error}
The $j$-th partial error is defined as 
$$
    \varepsilon_{-j} = Y - m_{-j}(X_{-j}).
$$ {#eq-partial-error}
::: 

Note that the tower property of conditional expectations imply the identity
$$
    m_{-j}(x_{-j}) = \mathbf{E}( m(X_j, x_{-j}) \mid X_{-j} = x_{-j}),
$$
which relate $m_{-j}$ to $m$. We can now show the following result on
implications of the hypothesis $H_0$ for the partial dependence 
function and the partial error. 

::: {#prp-dependence} 
If the regression function $m(x) = \mathbf{E}(Y \mid X = x)$ does not depend on $x_j$ then 

1. The partial dependence function $m_{j}$ is constant.
2. $m_{-j}(x_{-j}) = m(x_j, x_{-j})$ for any $x_j$.
3. $\mathrm{cov}(\varepsilon_{-j}, X_j) = 0$.
::: 

::: {.proof}
It follows directly from ([-@eq-partial-dependence]) that if $m$ does not depend on $x_j$ 
then $m_{j}(x_j) = \mathbf{E}(m(X_{-j}))$ is constant. 

To prove the second claim we note that if  $m$ does not depend on the $j$-th coordinate, 
$m(X_j, x_{-j}) = m(x_j, x_{-j})$ for any fixed $x_j$, whence 
$$
    m_{-j}(x_{-j}) = \mathbf{E}( m(X_j, x_{-j}) \mid X_{-j} = x_{-j}) =  m(x_j, x_{-j}).
$$

Finally, if $m$ does not depend on $x_j$, $\varepsilon_{-j} = \varepsilon = Y - \mathbf{E}(Y \mid X)$. 
Whence by the tower property of conditional expections, 
$$
    \begin{align*}
    \mathrm{cov}(\varepsilon_{-j}, X_j) & = \mathrm{cov}(\varepsilon, X_j) \\
    & = \mathbf{E}( \varepsilon X_j ) \\
    & = \mathbf{E}( \mathbf{E}(\varepsilon X_j \mid X)) \\
    & = \mathbf{E}( \mathbf{E}(\varepsilon \mid X) X_j) = 0,
    \end{align*}
$$
where we have used that $\mathbf{E}(\varepsilon \mid X) = 0$. 
::: 

There are various procedures, motivated by @prp-dependence, 
that are used in practice to better understand how $m$ depends on a 
particular predictor. To describe them, let $\hat{m}$ and $\hat{m}_{-j}$ 
denote estimates of $m$ and $m_{-j}$, respectively. 

1. The *partial dependence plot* is simply a plot of $\hat{m}_j(X_{i,j})$ against $X_{i,j}$, where 
    $$
        \hat{m}_j(X_{i,j}) = \frac{1}{n} \sum_{i'=1}^n \hat{m}(X_{i,j}, X_{i', -j}).
    $$
    It is a direct graphical check of whether $m_j$ is constant.
2.  For a $q$-dimensional model $\mathcal{M}$ of $m$ and a $q_0$ dimensional model $\mathcal{M}_{-j}$ of 
    $m_{-j}$, the $F$-test statistic is 
    $$
        F = \frac{\sum_{i=1}^n (\hat{m}_{-j}(X_{i,-j}) - \hat{m}(X_{i}))^2(q - q_0)}{\sum_{i=1}(Y_i - \hat{m}(X_{i}))^2 / (n - q)}.
    $$
    It quantifies directly any differences between $m_{-j}$ and $m$. 
3.  Defining the $j$-th *partial residuals* as 
    $$
        \hat{\varepsilon}_{i, -j} = Y_i - \hat{m}_{-j}(X_{-j}),
    $$ 
    we can plot $\hat{\varepsilon}_{i, -j}$ against $X_{i,j}$. 
4.  Letting $\hat{r}_j$ denote an estimate of the regression function 
    $$
        r_j(x_{-j}) = \mathbf{E}(X_j \mid X_{-j} = x_{-j}),
    $$ 
    we can also plot the partial residuals $\hat{\varepsilon}_{i,-j}$ against $X_{i,j} - \hat{r}_j(X_{i,-j})$.
    This plot is known as an *added variable plot*.

The suggested plots of the partial residuals against either the $j$-th predictor 
or its residual are both attempts to visualize dependencies between $\varepsilon_{-j}$ and $X_j$. 
The plots can be used to graphically detect if $\mathrm{cov}(\varepsilon_{-j}, X_j) = 0$ or not. 
We can also use the partial residuals to directly estimate the covariance, either as 
$$
    \tilde{\mathrm{cov}}_j = \frac{1}{n} \sum_{i=1}^n \hat{\varepsilon}_{i, -j} X_{i,j} 
$$ {#eq-partial-cov-bad}
or as 
$$
    \hat{\mathrm{cov}}_j = \frac{1}{n} \sum_{i=1}^n \hat{\varepsilon}_{i, -j} (X_{i,j} - \hat{r}(X_{i,-j})).
$$ {#eq-partial-cov-good}

Both of these covariance estimates quantify deviations from $H_0$, and in this way indirectly 
also how much $m$ depends on the $j$-th coordinate. The estimator ([-@eq-partial-cov-bad]) 
has the advantage that we do not need to estimate the regression function $r_j$, but it can
be severely biased and $\hat{\mathrm{cov}}_j$ is preferred in practice. 

To better appreciate the general definitions and procedures outlined above, we go through four 
different examples in increasing complexity, where most of the definitions and quantities 
above have some explicit expressions in terms of model parameters or components.

::: {#exm-linear-model}
If $m(x) = x^T \beta$ is linear we find that the partial dependence function is 
$$
    m_j(x_j) = \beta_j x_j + \mathbf{E}(X_{-j})^T \beta_{-j}.
$$ 
We see that the partial dependence function is affine with slope $\beta_j$. It 
is constant if and only if $\beta_j = 0$, which is the case if and only if $m$ 
does not depend on $x_j$. 

The $F$-test statistic defined above coincides for this model with the $F$-test statistic 
defined by @eq-F-test-statistic, and it can be used to formally test $H_0$. 

We also find that 
$$
    m_{-j}(x_{-j}) = x_{-j}^T \beta_{-j} + \beta_j \mathbf{E}(X_j \mid X_{-j} = x_{-j}) 
    = x_{-j}^T \beta_{-j} + \beta_j r_j(x_{-j}).
$$
Therefore 
$$
    m(x) - m_{-j}(x_{-j}) = \beta_j(x_j - r_j(x_{-j}))
$$
This difference is non-zero only if $\beta_j \neq 0$. Moreover, 
$$
    \varepsilon_{-j} = \varepsilon + \beta_j(X_j - r_j(X_{-j}))
$$
and we find that 
$$
    \mathrm{cov}(\varepsilon_{-j}, X_j) = \beta_j \mathbf{V}(X_j - r_j(X_{-j})).
$$
This covariance is thus proportional to $\beta_j$, with a proportionality 
constant $\mathbf{V}(X_j - r_j(X_{-j}))$ depending on the distribution of $X$ only. 
::: 

::: {#exm-partially-linear-model}
The regression function is called *partially linear* in the $j$-th variable if it 
is of the form 
$$
    m(x) = \beta_j x_j + h_{-j}(x_{-j}).
$$
In this case, similarly as above, the partial dependence function is an affine function
$$
    m_j(x_j) = \beta_j x_j + \mathbf{E}(h_{-j}(X_{-j})) 
$$
with slope $\beta_j$. Moreover, 
$$
    m_{-j}(x_{-j}) = h_{-j}(x_{-j}) + \beta_j \mathbf{E}(X_j \mid X_{-j} = x_{-j}) 
    = h_{-j}(x_{-j}) + \beta_j r_j(x_{-j}),
$$
and thus 
$$
    m(x) - m_{-j}(x_{-j}) = \beta_j(x_j - r_j(x_{-j}))
$$
exactly as if $m$ were linear. By the same argument as above, we get that 
$$
    \mathrm{cov}(\varepsilon_{-j}, X_j) = \beta_j \mathbf{V}(X_j - r_j(X_{-j})).
$$
::: 

The computations above for a partially linear regression function reveal that 
the partial dependence plot as well as procedures based on the partial residuals 
should reveal the same affine dependence of $m$ on $x_j$ whether $m$ 
is linear or only partially linear. The subtle, but important, 
practical difference is that this is only true if the estimators used 
for estimating the partial dependence function or for computing 
the partial residuals are flexible enough to correctly model the 
potentially nonlinear functions $h_{-j}$ and $m_{-j}$. 

::: {#exm-partially-linear-het-model}
If the regression function has the form
$$
    m(x) = \beta_j(x_{-j}) x_j + h_{-j}(x_{-j}),
$$
it is affine in $x_j$ for any fixed $x_{-j}$. It is
partially linear in the $j$-th variable but with a coefficient 
that depends on $x_{-j}$. The partial dependence function is 
the affine function
$$
    m_j(x_j) = \mathbf{E}(\beta_j(X_{-j})) x_j + \mathbf{E}(h_{-j}(X_{-j})) 
$$
with slope $\mathbf{E}(\beta_j(X_{-j}))$ being the mean slope of $x_j$ in $m$. 
$$
    m_{-j}(x_{-j}) = h_{-j}(x_{-j}) + \beta_j(x_{-j}) \mathbf{E}(X_j \mid X_{-j} = x_{-j}) 
    = h_{-j}(x_{-j}) + \beta_j(x_{-j}) r_j(x_{-j}),
$$
and thus 
$$
    m(x) - m_{-j}(x_{-j}) = \beta_j(x_{-j})(x_j - r_j(x_{-j})).
$$
The covariance becomes 
$$
    \begin{align*}
    \mathrm{cov}(\varepsilon_{-j}, X_j) & = \mathbf{E}(\beta_j(X_{-j})(X_j - r_j(X_{-j}))^2) \\
    & = \mathbf{E}(\beta_j(X_{-j})\mathbf{V}(X_j - r_j(X_{-j}) \mid X_{-j})) \\
    & = \mathbf{E}(\beta_j(X_{-j})v_j(X_{-j})), \\
    \end{align*}
$$
which we can interpret as a weighted mean of the individual slopes with weights 
$v_j(X_{-j}) = \mathbf{V}(X_j - r_j(X_{-j}) \mid X_{-j})$ being conditional variances. 
:::

::: {#exm-partially-additive-model}
The regression function is called *partially additive* in the $j$-th variable if 
it is of the form 
$$
    m(x) = m^0_j(x_j) + h_{-j}(x_{-j}).
$$
The partial dependence function is then 
$$
    m_j(x_j) = m^0_j(x_j) + \mathbf{E}(h_{-j}(X_{-j})), 
$$
which up to an additive constant coincides with $m^0_j$. 
The regression function $m_{-j}$ becomes 
$$
    m_{-j}(x_{-j}) = h_{-j}(x_{-j}) + \mathbf{E}(m^0(X_j) \mid X_{-j} = x_{-j}) 
    = h_{-j}(x_{-j}) + w_{-j}(x_{-j})
$$
where $w_{-j}(x_{-j}) = \mathbf{E}(m^0_j(X_j) \mid X_{-j} = x_{-j})$. Therefore 
$$
    m(x) - m_{-j}(x_{-j}) = m^0_j(x_j) - w_{-j}(x_{-j}). 
$$
Finally
$$
    \begin{align*}
        \mathrm{cov}(\varepsilon_{-j}, X_j) & =  \mathrm{cov}(m^0_j(X_j) - w_{-j}(X_{-j}), X_j) \\
        & = \mathbf{E}( \mathrm{cov}(m^0_j(X_j), X_j \mid X_{-j})).
    \end{align*}
$$
::: 

## Nonlinear basis expansions


### Nonlinear expansions for claims modeling {#sec-nonlinear-expansions}

We found in the earlier analysis of the insurance claims data a lack of
model fit. This could be due to a nonlinear relation between
`log(claims)` and `log(sum)`. To handle nonlinear
relations between the response and one or more predictors, the predictors
(as well as the response) can be nonlinearly transformed before they enter
into the linear model. Such pre-modeling transformations
extend the scope of the linear model considerably. For the insurance data
the original variables were already log-transformed, and
it is not obvious which other transformation to use.


```{r} 
#| label: fig-poly
#| fig-cap: "A basis of orthogonal polynomials."
#| column: margin
x <- seq(11, 21, 0.05)
pol <- as_tibble(cbind(x, poly(x, degree = 5)))
long_pol <- tidyr::pivot_longer(pol, cols = !x, names_to = "basis_fct")
p <- ggplot(NULL, aes(x, value, colour = basis_fct)) +
    scale_color_discrete(guide = "none") + xlab("") + ylab("")
p + geom_line(data = long_pol, linewidth = 1)
```


An alternative to data transformations is to use a small but
flexible class of *basis functions*, which
can capture the nonlinearity. One possibility is to use low degree polynomials.
This can be done by simply including powers of a predictor as additional
predictors.
[Due to the parsing of formulas in R,
powers have to be wrapped into an `I()`}-call, e.g., `y ~ x + I(x2)`.]
Depending on the scale and range of the predictor, this may
work just fine. However, raw powers can result in numerical difficulties.
An alternative is to use orthogonal polynomials, which are numerically more
well behaved.
\index{polynomials!orthogonal}

@fig-poly shows an example of an orthogonal basis of
degree 5 polynomials on the range 11--21. This corresponds approximately
to the range of `log(sum)`, which is a target for basis
expansion in the example. What should be noted in @fig-poly is
that the behavior near the boundary is quite erratic. This is characteristic
for polynomial bases. To achieve flexibility in the central part of the range,
the polynomials become erratic close to the boundaries. Extrapolation
beyond the boundaries cannot be trusted.

\index{spline}
An alternative to polynomials is splines. A spline[^splines]
is piecewisely a polynomial, and the pieces are joined together in a sufficiently smooth
way. The points where the polynomials are joined together are called
*knots*. The flexibility of a spline is determined by the number
and placement of the knots and the degree of the polynomials. A degree
$k$ spline is required to be $k-1$ times continuously differentiable.
A degree 3 spline, also known as a *cubic spline*, is a popular choice,
which thus has a continuous (piecewise linear) second derivative.

[^splines]:A spline is also a thin and flexible wood or metal strip used for smooth curve drawing.

```{r}
#| label: fig-b-spline
#| fig-cap: "A basis of cubic $B$-splines computed using the `bs()` function with two internal knots at 14 and 18 in addition to two boundary knots at 11 and 21."
#| column: margin
b_spline <- as_tibble(cbind(x, splines::bs(x, knots = c(14, 18))))
long_b_spline <- tidyr::pivot_longer(b_spline, cols = !x, names_to = "basis_fct")
p + geom_line(data = long_b_spline, linewidth = 1) +
    geom_rug(aes(x = c(11, 14, 18, 21), y = NULL, color = NULL), linewidth = 1)
```

\index{spline!cubic}
@fig-b-spline shows a basis of 5 cubic spline functions.
They are so-called $B$-splines \index{$B$-spline} (basis splines). Note that it is impossible
to visually detect the knots where the second derivative is
non-differentiable. The degree $k$ $B$-spline basis with $r$
internal knots has $k + r$ basis functions. The constant function is not included
then, and this is precisely what we want when the basis expansion is
used in a regression model including an intercept. As seen in @fig-b-spline,
the $B$-spline basis is also somewhat erratic close to the boundary. For a cubic
spline, the  behavior close to the boundary can be controlled by requiring that the second and
third derivatives are $0$ at the boundary knots. The result is known as a
*natural cubic spline*. \index{natural cubic spline} The extrapolation (as a spline) of a
natural cubic spline beyond the boundary knots is linear.

Due to the restriction on the derivatives of a natural cubic spline,
the basis with $r$ internal knots has $r + 1$ basis functions. Thus the basis for the
natural cubic splines with $r + 2$ internal knots has the same number of
basis functions as the raw cubic $B$-spline basis with $r$ internal knots. This means in
practice that, compared to using raw $B$-splines with $r$ internal knots, we can add two
internal knots, and thus increase the central flexibility of the model,
while retaining its complexity in terms of $r + 3$ parameters.

```{r}
#| label: fig-n-spline
#| fig-cap: "A $B$-spline basis for natural cubic splines computed using the `ns()` function with internal knots at 13, 15, 17 and 19 in addition to the two boundary knots at 11 and 21."
#| column: margin
n_spline <- as_tibble(cbind(x, splines::ns(x, knots = c(13, 15, 17, 19))))
long_n_spline <- tidyr::pivot_longer(n_spline, cols = !x, names_to = "basis_fct")
p + geom_line(data = long_n_spline, linewidth = 1) +
    geom_rug(aes(x = c(11, 13, 15, 17, 19, 21), y = NULL, color = NULL), linewidth = 1)
```

It is possible to use orthogonal polynomial expansions as well as spline expansions
in R together with the `lm()` function via the functions `poly()`,
`bs()` and `ns()`. The latter two are
in the splines package, which thus has to be loaded.

We will illustrate the usage of basis expansions for claim size modeling using
natural cubic splines. \index{spline expansion}

```{r}
#| label: claims-spline
#| code-fold: show
claims_lm_spline_add <- lm(
  log(claims) ~ ns(log(sum), knots = c(13, 15, 17, 19)) + grp,
  data = claims
)
claims_lm_spline_int <- lm(
  log(claims) ~ ns(log(sum), knots = c(13, 15, 17, 19)) * grp,
  data = claims
)
```

When we fit a model using basis expansions the coefficients of individual
basis functions are rarely interpretable or of interest. Thus reporting the
summary including all the estimated coefficients is not particularly informative.
Below we only extract the last lines from the summary to report the residual
variance and the $R^2$ quantities.

```{r, echo=2}
#| label: tbl-claims-spline-sum1
#| tbl-cap: "Summary statistics for the additive model based on a spline basis expansion."
#| tbl-cap-location: bottom
#| column: page-right
options(digits = 5)
glance(claims_lm_spline_add) |>
    knitr::kable()
options(digits = 7)
```

```{r, echo=2}
#| label: tbl-claims-spline-sum2
#| tbl-cap: "Summary statistics for the interaction model based on a spline basis expansion."
#| tbl-cap-location: bottom
#| column: page-right
options(digits = 5)
glance(claims_lm_spline_int) |>
    knitr::kable()
options(digits = 7)
```

Compared to the additive model without the basis expansion, $R^2$ (also the adjusted one)
is increased a little by the basis expansion. A further increase is observed for the model
that includes interactions as well as a basis expansion, and thus gives a unique nonlinear
relation between insurance sum and claim size for each trade group. @fig-claims-diag-spline
shows some diagnostic plots for the additive model, which show that the mean value model is
now pretty good, while the residuals are still slightly right skewed. The lack of normality
is not a problem for using the sampling distributions to draw inference. However, if we compute
prediction intervals, say, then we must account for the skewness in the residual distribution.
This can be done by using quantiles for the empirical distribution of the residuals instead of
quantiles for the normal distribution.

```{r}
#| label: fig-claims-diag-spline
#| message: false
#| fig-cap: "Residual plot (left), histogram (middle) of the raw residuals and qq-plot (right) of the standardized residuals for the additive model with a basis expansion of log insurance sum."
#| opts.label: widefigure3
#| cap-location: margin
claims_diag_spline <- augment(claims_lm_spline_add) ## Residuals etc.
gridExtra::grid.arrange(
  ggplot(claims_diag_spline, aes(.fitted, .resid), alpha = 0.2) + geom_point() + geom_smooth(),
  ggplot(claims_diag_spline, aes(.resid)) + geom_histogram(bins = 40),
  ggplot(claims_diag_spline, aes(sample = .std.resid)) + geom_qq() + geom_abline(),
  ncol = 3
)
```

Graphical comparisons are typically preferable over parameter comparisons when models involving nonlinear effects and basis expansions are compared. 

```{r}
#| label: fig-claims-scatter
#| message: false
#| code-fold: true
p0 <- ggplot(
        data = claims, 
        aes(sum, claims), 
        alpha = I(0.2)
    ) + 
    scale_x_log10(
        "Insurance sum (DKK)", 
        breaks = 10^c(6, 7, 8), 
        labels = c("1M", "10M", "100M")
        ) +
    scale_y_log10(
        "Claim size (DKK)", 
        breaks = 10^c(2, 4, 6, 8), 
        labels = c("100", "10K", "1M", "100M")
    ) +
    geom_point()
```

```{r}
#| label: fig-spline-fit
#| fig-cap: "Scatter plots and fitted values for the claim size models using natural cubic splines including 95% pointwise confidence bands. The additive model (blue) gives translations of the same nonlinear relation for all four trade groups wheras the interaction model (red) gives a separate nonlinear relation for all four trade groups."
#| opts.label: widefigure4
#| column: page
pred_add <- cbind(
    claims, 
    exp(predict(claims_lm_spline_add, interval = "confidence"))
)
pred_int <- cbind(
    claims, 
    exp(predict(claims_lm_spline_int, interval = "confidence"))
)
p0 <- p0 + facet_wrap(~ grp, ncol = 4) + coord_cartesian(ylim = c(100, 1e6))
p0 + geom_line(aes(y = fit), pred_add, color = "blue", linewidth = 2) +
  geom_ribbon(aes(ymax = upr, ymin = lwr), pred_add, fill = "blue", alpha = 0.2) +
  geom_line(aes(y = fit), pred_int, color = "red", linewidth = 2) +
  geom_ribbon(aes(ymax = upr, ymin = lwr), pred_int, fill = "red", alpha = 0.2)
```

@fig-spline-fit shows the fitted models
stratified according to trade group. The fitted values were computed using the `predict()`
function in R, which can also give upper and lower confidence bounds on the fitted
 values. The fitted value for the $i$-th observation is $X_i^T \hat{\beta}$. By 
@thm-pardisnorm a confidence interval for the fitted value can be
computed using @eq-confint with $a = X_i$. The
confidence intervals reported by `predict()` for lm objects are computed
using this formula, and these are the pointwise confidence bands shown in
@fig-spline-fit. From this figure any gain of the interaction model
over the additive model is questionable. The increase of $R^2$, say, appears to be mostly
driven by spurious fluctuations. For trade group 4 the capped
claims seem overly influential, and because trade group 3 has relatively few observations
the nonlinear fit is poorly determined for this group.

```{r}
#| label: claims-lm-add
#| code-fold: true
claims_lm_add <- lm(
    log(claims) ~ log(sum) + grp,
    data = claims
)
```

```{r}
#| label: tbl-claims-spline-anova
#| code-fold: show
#| tbl-cap: "Analysis of variance table comparing the two models based on spline expansions and the additive linear model."
#| tbl-cap-location: bottom
anova(
    claims_lm_add, 
    claims_lm_spline_add, 
    claims_lm_spline_int
) |> knitr::kable()
```

In the final comparison above using $F$-tests
we first test the hypothesis that the relation is linear in the additive model, and then we
test the additive hypothesis when the nonlinear relation is included.
The tests formally reject both hypotheses, but the
graphical summary in @fig-spline-fit suggests that the nonlinear additive model
is preferable anyway.

Introducing a penalty can dampen the spurious fluctuations the arose from
the nonlinear basis expansion. We conclude the analysis of the insurance data by 
illustrating how the penalized least squares fit can be computed, and what effect 
the penalty has on the least squares solution. First recall that the 
least squares estimator could have been computed by solving the normal equation 
directly. 

```{r}
#| label: LU
#| code-fold: show
X <- model.matrix(claims_lm_spline_int)
y <- model.response(model.frame(claims_lm_spline_int))
XtX <- crossprod(X)     # Faster than but equivalent to t(X) %*% X
Xty <- crossprod(X, y)  # Faster than but equivalent to t(X) %*% y
coef_hat <- solve(XtX, Xty)
# Comparison
range(coef_hat - coefficients(claims_lm_spline_int))
```

The result obtained by solving the normal equation directly is up to numerical errors 
identical to the solution computed using `lm()`. The R function `solve()` calls 
the Fortran routine DGESV from the LAPACK library 
for solution of linear equations using LU decomposition with partial pivoting (Gaussian 
elimination with row permutations). 

Likewise, we can compute the penalized estimator by solving the corresponding linear 
equation. We will do so where we only add a penalty on the coefficients corresponding to
the interaction terms in the nonlinear basis expansion. We will use penalty matrices of 
the form $\lambda \boldsymbol{\Omega}$ for a fixed matrix $\boldsymbol{\Omega}$ so 
that the amount of penalization is controlled by the parameter $\lambda > 0$. 

```{r}
#| label: penalize
#| code-fold: show
Omega <- diag(c(rep(0, 9), rep(1, 15)))
coef_hat <- cbind(
    solve(XtX + 0.1 * Omega, Xty),  # Small penalty
    solve(XtX + Omega, Xty),        # Medium penalty
    solve(XtX + 10 * Omega, Xty)    # Large penalty
)  
```

@fig-penalized shows the resulting penalized model compared to the 
unpenalized interaction model ($\lambda = 0$) and the additive model ($\lambda = \infty$). 
A minimal penalization is enough to
dampen the most pronounced spurious fluctuation for the fourth trade group 
considerably. 

```{r}
#| label: fig-penalized
#| fig-cap: "Scatter plots and fitted values for the claim size models where the nonlinear interaction terms have been penalized. The unpenalized interaction and additive model fits are also added."
#| opts.label: widefigure4
#| column: page
tmp <- cbind(
    coefficients(claims_lm_spline_int),
    coef_hat, 
    c(coefficients(claims_lm_spline_add), rep(0, 15))
)
pred <- cbind(claims[, c("sum", "grp")], exp(X %*% tmp)) 
pred <- tidyr::pivot_longer(pred, cols = !c(grp, sum), names_to = "lambda")                                 
p0 + geom_line(
  aes(y = value, color = lambda), 
  data = pred, 
  linewidth = 1) +
  theme(legend.position = "top") + 
  scale_color_discrete(
    expression(~lambda), 
    labels = c(0, 0.1, 1, 10, expression(infinity~~~(additive~model)))
  )     
```

The penalized least squares estimates can also be computed using the QR decomposition via
the function `lm.fit()`. To achieve this we need to augment the model matrix and the 
response data suitably. 

```{r}
#| label: fit-lamb
#| code-fold: show
fit_lamb1 <- lm.fit(rbind(X, sqrt(0.1) * Omega), c(y, rep(0, ncol(Omega))))
fit_lamb2 <- lm.fit(rbind(X, sqrt(10) * Omega),  c(y, rep(0, ncol(Omega))))
## Comparisons
range(coef_hat[, 1] - coefficients(fit_lamb1))  
range(coef_hat[, 3] - coefficients(fit_lamb2))  
```

The results are again (numerically) identical to using `solve()` above. 
If several penalized estimates using matrices $\lambda \boldsymbol{\Omega}$ for different choices 
of $\lambda$ are to be computed, a third option using a diagonalization of $\mathbf{X}^T \mathbf{X}$ is beneficial. 
This is implemented in, e.g., the `lm.ridge()` function in the package MASS, which unfortunately 
doesn't let you choose your own penalty matrix -- it only implements $\boldsymbol{\Omega} = \mathbf{I}$.

In conclusion, we have found evidence in the data for an overall nonlinear
relation between insurance sum and claim size plus an additive effect of trade group. For small insurance sums (less than DKK 1M) there is almost no relation, while for larger insurance
sums there is roughly a linear relation on a log-log scale. The model fits the data
well with a non-normal and slightly right skewed residual distribution, but as a
predictive model it is weak, and insurance sum can together with
trade group only explain 6-7\% of the variation in the (log) claim size distribution.


### Nonlinear birth weight models  {#sec-nonlinear-birth}

Basis expansions can be tried if we expect to
get something out of it, that is, if we expect that there are some
nonlinear relations in the data. On the other hand, if we construct
models tailor-made to capture
nonlinear relations we have spotted by eye-balling residual plots we
run the risk of overfitting random fluctuations. In particular,
formal downstream justifications using statistical tests are invalidated.
The eye-balling process is a model selection procedure with statistical
implications, which are difficult to account for.

We decided to expand `gestationalAge` using natural cubic splines
with three knots in 38, 40, and 42 weeks. The boundary knots were determined
by the range of the data set, and were thus 25 and 47. We also expanded
`age`, but we let the `ns()` function
determine the knots automatically for that predictor. The last continuous
predictor, `alcohol` has a very skewed marginal distribution, and it
was not judged to be suitable for a standard basis expansion.
We also present a test of the nonlinear effect. \index{spline expansion}

```{r}
#| label: filter-birth-weight-again-and-lm
data("birth_weight")
birth_weight <- filter(
  birth_weight,
  weight > 32,
  length > 10 & length < 99,
  gestationalAge > 18
) |> 
  na.omit() |>
  select(- c(interviewWeek, fetalDeath))

form <- weight ~ gestationalAge + 
                 age + 
                 children +
                 coffee + 
                 alcohol + 
                 smoking + 
                 abortions + 
                 feverEpisodes

birth_weight_lm <- lm(form, data = birth_weight)


binScale <- scale_fill_continuous(
  breaks = c(1, 10, 100, 1000),
  low = "gray80", 
  high = "black",
  trans = "log", 
  guide = "none"
)
```

[Nonlinear main effects model.]{.aside}

```{r}
#| label: tbl-test-nonlinear-model
#| tbl-cap: "Test of the model including a spline expansion of `gestationalAge` against the main effects model."
#| tbl-cap-location: bottom
#| code-fold: show
nsg <- function(x) ns(x, knots = c(38, 40, 42), Boundary.knots = c(25, 47))

form <- weight ~ nsg(gestationalAge) + 
    ns(age, df = 3) + 
    children +
    coffee + 
    alcohol + 
    smoking + 
    abortions + 
    feverEpisodes

birth_weight_lm_spline <- lm(form, data = birth_weight)

anova(birth_weight_lm, birth_weight_lm_spline) |> knitr::kable()
```

The main effects model is a submodel of the nonlinear main effects model.
This is not obvious. The nonlinear expansion will result in 4 columns in
the model matrix $\mathbf{X}$, none of which being `gestationalAge`. However, the
linear function is definitely a natural cubic spline, and it is thus in the
span of the 4 basis functions. With $\mathbf{X}'$ the model matrix for the main
effects model, it follows that there is a $C$ such that @eq-subspace holds.
This justifies the use of the $F$-test. The conclusion from @tbl-test-nonlinear-model
is that the nonlinear model is highly significant.


```{r}
#| label: fig-spline-model-diag
#| fig-cap: "Diagnostic plots for the model with `gestationalAge` expanded using splines."
#| fig-cap-location: margin
#| opts.label: widefigure3
#| message: false
birth_weight_diag <- augment(birth_weight_lm_spline, data = birth_weight)
p1 <- ggplot(birth_weight_diag, aes(.fitted, .std.resid)) +
  stat_binhex(bins = 20) + binScale + geom_smooth(linewidth = 1, fill = "blue") +
  xlab("fitted values") + ylab("standardized residuals")
p2 <- ggplot(birth_weight_diag, aes(gestationalAge, .std.resid)) +
  stat_binhex(bins = 20) + binScale + geom_smooth(linewidth = 1, fill = "blue") +
  xlab("gestationalAge") + ylab("")
p3 <- ggplot(birth_weight_diag, aes(sample = .std.resid)) +
  geom_abline(intercept = 0, slope = 1, color = "blue", linewidth =  1) +
  geom_qq() + 
  xlab("theoretical quantiles") + ylab("")
gridExtra::grid.arrange(p1, p2, p3, ncol = 3)
```

The positioning of the knots for spline expansion
is selected based on the marginal distribution of the predictor[^knots]
The knots should be placed reasonably relative to the distribution of the
predictor, so that we learn about nonlinearities where there is data to learn from.
In this case we placed knots at the median and the 10% and 90% quantiles
of the distribution of `gestationalAge`. The `ns()` function
makes a similar automatic selection of knots based on the marginal distribution
of the predictor variables when it is applied in the formula. One just have to
specify the number of basis functions using the `df` argument.


[^knots]: Letting the knots be parameters to be estimated is not a statistically viable idea. The resulting estimation problem becomes computationally much more difficult, and it is not straightforward how to adjust subsequent statistical analyses to account for the data adaptive choice of knots.

An eye-ball decision based on
@fig-weight-model-control would be that a single knot around 41 would have done the job,
but we refrained from making such a decision. A subsequent test of the
nonlinear effect with 1 degrees of freedom would not appropriately take
into account how the placement of the knot was made. 


@fig-spline-model-diag shows diagnostic plots for the nonlinear main
effects model. They show that the inclusion of the nonlinear effect
removed the previously observed problem with assumption **A1**. The error distribution
is still not normal, but right skewed with a fatter right tail than the normal
distribution. There is a group
of extreme residuals for preterm born children, which should be given more attention
than they will be given here.


[Reduced nonlinear main effects model.]{.aside}

```{r}
#| label: tbl-spline-small
#| tbl-cap: "Test of weak predictors in the nonlinear main effects model."
#| tbl-cap-location: bottom
#| code-fold: show
form <- weight ~ nsg(gestationalAge) + 
    children + 
    coffee + 
    smoking

birth_weight_lm_spline_small <- lm(form, data = birth_weight)

anova(birth_weight_lm_spline_small, birth_weight_lm_spline) |> 
  knitr::kable()
```


@tbl-spline-small shows that dropping the four weak predictors from
the nonlinear main effects model is again borderline and not really significant.

@fig-spline-fit-birth shows examples of the fit for the reduced nonlinear main effects model.
The figure illustrates the general nonlinear relation between `weight` and `gestationalAge`. Differences
due to other variables are purely additive in this model, which amounts to translations up
or down of the curve. The figure shows a couple of extreme cases; the majority group who
have had children before and who do not smoke or drink coffee, and a minority group who have
had children before and smoke and drink coffee the most. What we should notice is
the wider confidence band on the latter (`smoke = 3`, `coffee = 3`) compared
to the former, which is explained by the skewness of the predictor distributions.
@tbl-spline-conf-int gives confidence intervals for the remaining parameters
based on the nonlinear main effects model.

```{r}
#| label: fig-spline-fit-birth
#| fig-cap: "Main effects model with basis expansion of `gestationalAge`. Here illustrations of the fitted mean and 95% confidence bands for `children=1`."
#| fig-cap-location: margin
#| opts.label: widefigure2
pred_frame <- expand.grid(
  children = factor(1),
  smoking = factor(c(1, 3)),
  coffee = factor(c(1, 3)),
  gestationalAge = seq(25, 47, 0.1),
  alcohol = 0,
  age = median(birth_weight$age),
  feverEpisodes = 0,
  abortions = factor(0)
)

pred <- predict(
  birth_weight_lm_spline, 
  newdata = pred_frame,
  interval = "confidence"
)

pred_frame <- cbind(pred_frame, pred)

ggplot(pred_frame, aes(gestationalAge, fit, color = coffee)) + 
  geom_line() + ylab("weight") +
  geom_ribbon(
    aes(ymin = lwr, ymax = upr, fill = coffee), 
    alpha = 0.3
  ) + facet_grid(. ~ smoking, label = label_both)
```

```{r}
#| label: tbl-spline-conf-int
#| tbl-cap: "Confidence intervals."
#| tbl-cap-location: bottom
#| column: margin
#| echo: false
tidy(birth_weight_lm_spline, conf.int = TRUE)[9:14, ] |>
  select(c(term, conf.low, conf.high)) |> knitr::kable(digits = 3)
```

To conclude the analysis, we will again include interactions. However, the variable
`gestationalAge` is in fact only taking the integer values 25 to 47,
and the result of a nonlinear effect coupled with a third order interaction, say,
results in an almost saturated model. That is, the interaction model has more or less a separate mean for
all observed combinations of the predictors. We choose to consider a less complex model with all
second order interactions between `gestationalAge` and the three factors that
we have judged to be strong predictors.

[Nonlinear interaction model.]{.aside}

```{r}
#| label: nonlinear-int-model
#| code-fold: show
form <- weight ~ (smoking + coffee + children) * nsg(gestationalAge) +
   ns(age, df = 3) + 
   alcohol + 
   abortions + 
   feverEpisodes

birth_weight_lm_spline_int <- lm(form, data = birth_weight)
```


```{r}
#| label: fig-spline-int-fit
#| fig-cap: "Comparison of the interaction model (red, 95% gray confidence bands) with the nonlinear main effects model (blue)."
#| opts.label: widefigure2
#| column: page
pred_frame <- expand.grid(
  children = factor(c(0, 1)),
  smoking = factor(c(1, 2, 3)),
  coffee = factor(c(1, 2, 3)),
  gestationalAge = 25:47,
  alcohol = 0,
  age = median(birth_weight$age),
  feverEpisodes = 0,
  abortions = factor(0)
)

pred <- predict(birth_weight_lm_spline_int, newdata = pred_frame, interval = "confidence")


pred_frame <- cbind(pred_frame, pred)
pred_frame$fit_spline <- predict(birth_weight_lm_spline, newdata = pred_frame)

ggplot(pred_frame, aes(gestationalAge, fit)) +
  geom_point(data = birth_weight, aes(y = weight), size = 0.3, alpha = 0.1) + 
  facet_grid(coffee ~ children + smoking, label = label_both) +
  geom_ribbon(aes(ymin = lwr, ymax = upr), fill = gray(0.85)) +
  geom_line(color = "red") + coord_cartesian(ylim = c(0, 5000)) +
  geom_line(aes(y = fit_spline), color = "blue") + ylab("weight") +
  scale_y_continuous(breaks = c(1000, 2000, 3000, 4000))
```


@fig-spline-int-fit shows the predicted values for the reduced nonlinear main effects model for
all combinations of `children`, `smoking` and `coffee` (blue curves).
These curves are all just translations of each other. In addition, the figure
shows the predicted values and a 95% confidence band as estimated with the
nonlinear interaction model. We observe minor deviations for the
reduced main effects model, which can explain the significance of the
test, but the deviations appear unsystematic and mostly related to
extreme values of `gestationalAge`. The conclusion is that even
though the inclusion of interaction effects is significant, there is
little to gain over the reduced nonlinear main effects model.

```{r}
#| label: tbl-nonlinear-int-model
#| tbl-cap: "Test of nonlinear interaction model."
#| tbl-cap-location: bottom
anova(birth_weight_lm_spline, birth_weight_lm_spline_int) |> 
  knitr::kable(digits = 10)
```

@tbl-nonlinear-int-model shows that inclusion of interaction terms is still significant by a
formal statistical test. It is, however, always a good idea to visualize how the nonlinear interaction model
differs from the model with only main effects as above instead of just focusing on the formal test. The test
may, on the other hand, support the visualization as a quantification of whether the differences
seen in a figure are actually statistically significant.

## Practical data analysis

```{r}
#| eval: false
#| echo: false
library(neuralnet)
standard <- function(x) (x - min(x)) / (max(x) - min(x))
unstandard <- function(x, min, max) x * (max - min) + min 

birth_weight_sub <- select(birth_weight, c(weight, gestationalAge)) |> 
    mutate(weight_st = standard(weight), gestationalAge_st = standard(gestationalAge)) 
softplus <- function(x) log(1 + exp(x))
form <- weight_st ~ gestationalAge_st ## + age + smoking + coffee + children + alcohol + abortions + feverEpisodes
tmp <- neuralnet(form, data = birth_weight_sub, hidden = 2, linear.output = FALSE, act.fct = softplus)
```

```{r}
#| eval: false
#| echo: false
birth_weight_sub <- mutate(
    birth_weight_sub, 
    pred = unstandard(predict(tmp, newdata = birth_weight_sub)[, 1], min(weight), max(weight)) 
)
plot(tmp)
```

```{r}
#| eval: false
#| echo: false
ggplot(birth_weight_sub, aes(gestationalAge, weight)) + 
    geom_point(size = 1, alpha = 0.1) +
    geom_line(aes(y = pred), color = "red")
```

```{r}
#| eval: false
#| echo: false
library(keras3)
x_train <- birth_weight$gestationalAge |> as.matrix()
y_train <- birth_weight$weight
# x_train <- standard(x_train)
y_train <- standard(y_train)
```

```{r}
#| eval: false
#| echo: false
model <- keras_model_sequential(input_shape = 1)
normalizer <-  layer_normalization() 
normalizer |>  adapt(x_train)
model |>
    normalizer() |>
    layer_dense(units = 100, activation = "relu") |>
    layer_dense(units = 100, activation = "relu") |>
    layer_dense(units = 1)
```

```{r}
#| eval: false
#| echo: false
compile(
    model, 
    loss = "mse"
)
```

```{r}
#| eval: false
#| echo: false
# x_train_noise <- x_train #+ rnorm(length(x_train), 0, 0.05)
history <- fit(model, x_train, y_train, batch_size = 1000, epochs = 100)
```

```{r}
#| eval: false
#| echo: false
plot(history)
plot(x_train, y_train)
lines(sort(x_train), predict(model, sort(x_train)), type = "l", ylim = c(0, 1), col = "blue")
```


## Causal interpretations

```{r}
#| eval: false
#| echo: false
c(29, 40, 50, 52, 54, 56, 58, 63) # Full range
# The removes the smallest and largest
birth_weight_mut <- mutate(birth_weight, length_bin = cut(length, breaks = c(40, 50, 52, 54, 56, 58))) |>
    na.omit()

ggplot(birth_weight_mut, aes(coffee, weight, fill = length_bin)) + 
    geom_boxplot(outliers = FALSE) 
```

```{r}
#| eval: false
#| echo: false
birth_weight_sum <- group_by(birth_weight_mut, coffee) |>
    summarize(mean_weight = mean(weight))

group_by(birth_weight_mut, length_bin, coffee) |>
    summarize(mean_weight = mean(weight), freq = n()) |>
    ungroup() |>
    group_by(coffee) |>
    mutate(freq = freq / sum(freq)) |> 
    ggplot(aes(coffee, mean_weight, color = length_bin, group = length_bin)) + 
    geom_line() + 
    geom_point() + 
    geom_point(data = birth_weight_sum, aes(color = NULL, group = 1)) +
    geom_line(data = birth_weight_sum, aes(color = NULL, group = 1))
```

```{r}
#| eval: false
#| echo: false
lm(weight ~ coffee + length, data = birth_weight) |> tidy()
lm(weight ~ coffee , data = birth_weight) |> tidy()
```

```{r}
#| eval: false
#| echo: false
ggplot(birth_weight_mut, aes(gestationalAge, weight, color = length)) + 
    geom_point() + 
    geom_smooth(aes(group = length_bin), color = "black", linetype = 2, method = "lm", se = FALSE) + 
    geom_smooth(aes(color = NULL, group = NULL), color = "black", method = "lm", se = FALSE) + 
    scale_color_fermenter(palette = "GnBu", breaks = c(40, 50, 52, 54, 56, 58))
```

```{r}
#| eval: false
#| echo: false
lm(weight ~ gestationalAge * length, data = birth_weight) |> tidy()
lm(weight ~ gestationalAge + length, data = birth_weight) |> tidy()
lm(weight ~ gestationalAge, data = birth_weight) |> tidy()
```


```{r}
#| label: a-tikz-figure
#| engine: tikz
#| eval: false
#| echo: false
\begin{tikzpicture}[description/.style={fill=white,inner sep=2pt}] 
\matrix (m) [matrix of math nodes, row sep=3em, column sep=3em, text height=1.5ex, text depth=0.25ex] {
 \textrm{Low birth weight} & \textrm{Risk factors} \\
 \textrm{Smoking} & \textrm{Mortality}\\}; 
\path[->] 
(m-1-2) edge (m-1-1) 
(m-2-1) edge (m-1-1)
(m-1-2) edge (m-2-2);
\end{tikzpicture}
```


```{r}
#| label: another-tikz-figure
#| engine: tikz
#| column: page
#| eval: false
#| echo: false
\begin{tikzpicture}

  \draw[->] (-0.5,0) -- (5.5,0) node[right] {predictors}; 
  \draw[->] (0,-0.5) -- (0,4) node[above] {$|\beta_i|/\text{se}_i$}; 
  \draw[-, blue!40!white] (0.4, 0) node [below, black] {1} -- (0.4, 3);
  \draw[-, blue!40!white] (0.8, 0) -- (0.8, 2.9);
  \draw[-, blue!40!white] (1.2, 0) -- (1.2, 2.8);
  \draw[-, blue!40!white] (1.6, 0) -- (1.6, 2.4);
  \draw[-, blue!40!white] (2.0, 0) -- (2.0, 2);
  \draw[-, blue!40!white] (2.4, 0) -- (2.4, 1.5);
  \draw[-, blue!40!white] (2.8, 0) -- (2.8, 1.3);
  \draw[-, blue!40!white] (3.2, 0) -- (3.2, 1);
  \draw[-, blue!40!white] (3.6, 0) -- (3.6, 0.8);
  \draw[-, blue!40!white] (4.0, 0) -- (4.0, 0.5);
  \draw[-, blue!40!white] (4.4, 0) -- (4.4, 0.4);
  \draw[-, blue!40!white] (4.8, 0) node [below, black] {$p$} -- (4.8, 0.2);
  \draw[-, red!40!white] (-0.1, 1)  node [left] {2} -- (5, 1);
  \node at (5, 4) {(A)};

  \begin{scope}[shift = {(0, 5)}]
  \draw[->] (-0.5,0) -- (5.5,0) node[right] {predictors}; 
  \draw[->] (0,-0.5) -- (0,4) node[above] {$|\beta_i|/\text{se}_i$}; 
  \draw[-, blue!40!white] (0.4, 0) node [below, black] {1} -- (0.4, 3);
  \draw[-, blue!40!white] (0.8, 0) -- (0.8, 2.9);
  \draw[-, blue!40!white] (1.2, 0) -- (1.2, 2.8);
  \draw[-, blue!40!white] (1.6, 0) -- (1.6, 2.4);
  \draw[-, blue!40!white] (2.0, 0) -- (2.0, 2);
  \draw[-, blue!40!white] (2.4, 0) -- (2.4, 1.5);
  \draw[-, blue!40!white] (2.8, 0) -- (2.8, 1.3);
  \draw[-, blue!40!white] (3.2, 0) -- (3.2, 1);
  \draw[-, blue!40!white] (3.6, 0) -- (3.6, 0.8);
  \draw[-, blue!40!white] (4.0, 0) -- (4.0, 0.5);
  \draw[-, blue!40!white] (4.4, 0) -- (4.4, 0.4);
  \draw[-, blue!40!white] (4.8, 0) node [below, black] {$p$} -- (4.8, 0.2);
  \draw[-, red!40!white] (-0.1, 1)  node [left] {2} -- (5, 1);
  \node at (5, 4) {(A)};
  \end{scope}
\end{tikzpicture}
```


## Exercises {.unnumbered}

::: {#exr-formulas-spline-expansions}

This is a continuation of Exercise @exr-formulas.

```{r}
#| label: formula-sim-data
#| code-fold: show
set.seed(04022016)
factor_labels <- rep(c("A", "B", "C"), each = 50)
formula_data <- tibble(
    x =  rnorm(150, 0),
    z = factor(sample(factor_labels)),
    mu = (z == "A") + cos(x) + 0.5 * (z == "B") * sin(x),
    y = rnorm(150, mu, 0.5)
)
```


Expansions can be achieved by applying an R function that returns a 
matrix of basis function evaluations when applied to a vector of 
predictor values. We consider a spline basis expansion.

```{r}
#| label: formula-expansions
#| code-fold: show
library(splines)
model.matrix(y ~ ns(x, df = 4), formula_data) |>
    head()
```

Before you run the R expression below, answer the following question on the 
basis of what you have learned.

::: 

1.  Determine the number of columns for the model matrix below and 
    explain and interpret what the columns mean?

```{r}
#| label: formula-expansions2
#| eval: false
#| code-fold: show
model.matrix(y ~ ns(x, df = 4) * z, formula_data) |>
    head()
```

@fig-formula-linear-model was obtained by fitting the linear model using the model 
matrix, or rather the formula, from above. The figure illustrates the model fit by 
comparing the fitted model directly to the true means and via the residuals. 
Fitted values and residuals were computed using the `augment()` function from the 
broom package. 

```{r}
#| label: fig-formula-linear-model
#| opts.label: widefigure
#| echo: false
#| message: false
#| fig-cap: "Left: The fitted values plotted as lines together with the data. The dashed lines are the true means. Right: a residual plot."
#| cap-location: margin
y_lm <- lm(y ~ ns(x, df = 4) * z, formula_data)
formula_data_aug <- broom::augment(y_lm, data = formula_data)
p1 <- ggplot(formula_data_aug, aes(x, y, color = z)) + 
    geom_point() + geom_line(aes(y = mu), linetype = 2) + 
    geom_line(aes(y = .fitted)) + theme(legend.position = "top")
p2 <- ggplot(formula_data_aug, aes(.fitted, .resid, color = z)) + 
    geom_point() + theme(legend.position = "top", legend.box.background = element_rect(fill = NA, color = NA)) + 
    geom_smooth(aes(color = NULL), color = NA) +
    geom_smooth(aes(color = NULL), se = FALSE)
gridExtra::grid.arrange(p1, p2, ncol = 2) 
```

2. Reproduce @fig-formula-linear-model and discuss the model fit.


::: {#exr-column-space-bspline}
Consider the following matrices

```{r}
#| code-fold: false
X <- splines::bs(
    seq(0, 1, 0.01), 
    knots = c(0.3, 0.5, 0.7)
)
X_prime <- seq(0, 1, 0.01)
```

of $B$-spline basis functions evaluated on a grid in the interval $[0, 1]$. 
Use @exr-column-space to verify that `X_prime` is in the
column space of `X`. You can use the formula from 
the exercise to compute $\mathbf{C}.$ Can you also use `lm.fit()`?
:::